---
title: "Pseudotime plotting"
output: html_notebook
---

Zoë Steier

* Pseudotime plots
* In-silico gating

# Load packages

```{r Load packages}
library(tidyverse)
library(scales)
library(RColorBrewer)
library(pROC)
library(slingshot)

packageVersion("tidyverse")
packageVersion("scales")
packageVersion("RColorBrewer")
packageVersion("pROC")
packageVersion("slingshot")
```

# Load data

```{r Load totalVI data}
totalvi_path <- "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/totalVI_thymus/thymus111_allbatches_stable_posselecting/totalVI_results/"

latent <- read_csv(gzfile(str_c(totalvi_path, "latent.csv.gz")))
umap_totalVI <- read_csv(gzfile(str_c(totalvi_path, "umap.csv.gz")))

denoised_proteins <- read_csv(gzfile(str_c(totalvi_path, "denoised_proteins.csv.gz")))
denoised_genes <- read_csv(gzfile(str_c(totalvi_path, "denoised_genes.csv.gz")))

totalVI_genes <- read_csv(gzfile(str_c(totalvi_path, "totalVI_genes.csv.gz")))
totalVI_proteins <- read_csv(gzfile(str_c(totalvi_path, "totalVI_proteins.csv.gz")))

#raw_proteins <- read_csv(gzfile(str_c(totalvi_path, "raw_proteins.csv.gz")))
#raw_genes <- read_csv(gzfile(str_c(totalvi_path, "raw_genes.csv.gz")))

obs <- read_csv(gzfile(str_c(totalvi_path, "obs.csv.gz")))

meta_csv <- read_csv("/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/metadata/Metadata_SeqStats_totalSeq_experiments.csv")
meta_thymus <- meta_csv %>%
  filter(Tissue == "thymus") %>%
  filter(TotalSeq_panel == "ADT111") 
meta_thymus <- meta_thymus %>%
  mutate(Batch = seq(1, dim(meta_thymus)[1]) - 1) # rename batches starting at 0 (19 batches in total)
```

```{r Load filtered data}
filt_path <- "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/Annotation_thymus/"

obs_filtered <- read_csv(gzfile(str_c(filt_path, "adata_positiveselection_filt_obs.csv.gz")))

filtered_cells <- obs_filtered$X1
# length(unique(filtered_cells)) # 29408
```

```{r Parse totalVI data}
# Extract: clusters, UMAP, denoised proteins, latent, etc.
# Remove first column of dataframes - this is an index column coming from python

# totalVI UMAP
totalvi_umap <- as.data.frame(umap_totalVI[, -1])
colnames(totalvi_umap) <- c("UMAP1", "UMAP2")
row.names(totalvi_umap) <- obs$X1

# totalVI latent space
totalvi_latent <- as.matrix(latent[, -1])
row.names(totalvi_latent) <- obs$X1

# totalVI denoised proteins
denoised_proteins <- denoised_proteins %>%
  select(-c(X1))
colnames(denoised_proteins) <- totalVI_proteins[, 2][[1]]
denoised_proteins <- denoised_proteins %>%
  mutate(Barcode = obs$X1)

# totalVI denoised genes
denoised_genes <- denoised_genes %>%
  select(-c(X1))
colnames(denoised_genes) <- totalVI_genes[, 2][[1]]
denoised_genes <- denoised_genes %>%
  mutate(Barcode = obs$X1)

# # Raw genes
# raw_genes <- raw_genes %>%
#   select(-c(X1))
# colnames(raw_genes) <- totalVI_genes[, 2][[1]]
# raw_genes <- raw_genes %>%
#   mutate(Barcode = obs$X1)
# 
# # Raw proteins
# raw_proteins <- raw_proteins %>%
#   select(-c(X1))
# colnames(raw_proteins) <- totalVI_proteins[, 2][[1]]
# raw_proteins <- raw_proteins %>%
#   mutate(Barcode = obs$X1)
```

```{r Make metadata from totalVI}
# totalVI Leiden clusters and metadata
totalVI_meta <- obs %>%
  select(Batch = 'batch_indices', UMIs_RNA = "n_counts", UMIs_protein = "n_protein_counts",
          'n_genes', 'n_proteins', 'percent_mito', contains("leiden_totalVI_"), Barcode = "X1") %>%
  mutate_at(vars(contains("leiden_totalVI_")), as_factor) %>% # Convert cluster labels to factors
  left_join(meta_thymus %>% select(Sample, Genotype, Experiment, Replicate, Location, Batch), by = "Batch") %>% # Join with experimental metadata
  mutate(Mouse = str_c(Genotype, "_", Replicate)) %>% # Add mouse information
  mutate_at(c("Sample", "Batch", "Mouse", "Experiment", "Genotype", "Location"), as_factor) %>% # Set as factors to preserve the order of levels
  select(-"Replicate")
```

```{r Load pseudotime data}
pseudotime_data <- read_csv("pseudotime_slingshot_2020913.csv", col_types = cols(Lineage_by_genotype = col_character(), 
                                                                                  Lineage_by_genotypeSlingshot = col_character()))
time_bins <- pseudotime_data %>%
  mutate(Pseudotime_bin = cut(mean_pseudotime, breaks = seq(0,16, by = 2), include.lowest = TRUE,
                              labels = str_c(as.character(seq(0, 7)))))

# Assign cells with pseudotime > 16 to the last bin
time_bins[is.na(time_bins$Pseudotime_bin), "Pseudotime_bin"] <- "7" # This is exactly 20 cells

pseudotime_data
```

# Plot filtering UMAP

```{r Label clusters to remove}
colnames(obs)[1] <- "Barcode"
colnames(obs_filtered)[1] <- "Barcode"

cd8_outliers <- obs$Barcode[!(obs$Barcode %in% obs_filtered$Barcode)]


filtering_labels <- as_tibble(totalvi_umap, rownames = "Barcode") %>%
  left_join(obs, by = "Barcode") %>%
  mutate(Filtered_labels = factor(case_when(leiden_totalVI_res1.4 == "22" ~ "Interferon sig.",
                                     leiden_totalVI_res1.4 == "19" ~ "Doublet",
                                     leiden_totalVI_res1.4 == "16" ~ "Cycling",
                                     leiden_totalVI_res1.4 == "21" ~ "GD-like",
                                     leiden_totalVI_res1.4 == "5" ~ "Neg. sel. (2)",
                                     leiden_totalVI_res0.6 == "6" ~ "Treg",
                                     leiden_totalVI_res0.6 == "8" ~ "Interferon sig.",
                                     leiden_totalVI_res0.6 == "9" ~ "Doublet",
                                     Barcode %in% cd8_outliers ~ "CD8tg outliers",
                                     TRUE ~ "Positively-selecting"), levels = c("CD8tg outliers", "Cycling", "Doublet", "GD-like", "Interferon sig.", "Neg. sel. (2)",
                                                                              "Treg", "Positively-selecting")))
```

```{r Plot labeled cells}
ggplot(filtering_labels, aes(x = UMAP1, y = UMAP2, color = Filtered_labels)) +
  geom_point(stroke = 0, size = .4) +
  scale_color_manual(values = c(hue_pal()(7), "gray50")) +
  theme_classic() +
  labs(color = "") +
  guides(colour = guide_legend(override.aes = list(size=2))) +
  coord_equal()

ggsave("/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/Annotation_thymus/figures/annotate_positiveselection_filter.png", 
       bg = "transparent", height = 4, width = 6, dpi = 1000) 
```

# Pseudotime plotting
## Prepare data

```{r Pseudotime metadata}
pseudotime_data <- time_bins %>% # pseudotime_data, but now add in time bins
  mutate(Lineage_by_genotypeWT = Lineage_by_genotype) %>%
  mutate(Lineage_by_genotypeWT = case_when(is.na(Lineage_by_genotypeWT) ~ "WT",
                                           !(is.na(Lineage_by_genotypeWT)) ~ Lineage_by_genotypeWT))

# Clean up the lineages by removing the "wrong lineage cells" and keeping the WT cells
pseudotime_data <- pseudotime_data %>%
  mutate(Lineage_by_genotypeSlingshotWT = Lineage_by_genotypeSlingshot) %>%
  mutate(Lineage_by_genotypeSlingshotWT = case_when(Lineage_by_genotypeWT == "WT" ~ "WT", 
                                           !(Lineage_by_genotypeWT == "WT") ~ Lineage_by_genotypeSlingshot)) # if it is not WT, it is either CD4, CD8, or NA

table(pseudotime_data$Lineage_by_genotypeSlingshotWT)
#sum(is.na(pseudotime_data$Lineage_by_genotypeSlingshotWT)) # 1050 wrong lineage cells
```

```{r Merge genes and meta}
denoised_annotated <- denoised_genes %>%
  left_join(pseudotime_data %>% select(Barcode, mean_pseudotime, Lineage_by_genotypeSlingshotWT, Genotype, Mouse), by = "Barcode") %>% # add mouse for revision 2
  filter(Lineage_by_genotypeSlingshotWT %in% c("CD4", "CD8", "WT"))
```

```{r Scale genes}
genes_to_plot <- c("Rag1", "Rag2", "Cxcr4", "Ccr7", "Runx3", "Zbtb7b", "Cd4", "Cd8a", "Cd8b1", "Trbc1", "Ccr4", "Ccr9", "Cd24a", "H2-K1", "Klf2", "S1pr1", "Cd55", "Cd5", "Cd69", "Bcl2", "Sell", "Gata3", "Il7r", 
                   "Id2", "Id3", "Cd40lg", "Dusp2", "Nr4a1", "Kcna2", "Tmie", "Themis", "Dusp5",
                   "Nfatc2", "Nfatc1", "Nfatc3", "Nfat5", "Egr1", "Egr2", "Egr3", "Ets1", "Stat5a", "Stat5b",
                   "Tcf3", "Tcf12",
                   "Batf", "Cd28", "Lef1", "Mapk11", "Ikzf2", "Nfkb2", "Relb",
                   "Tesc",
                   "Ndrg1", "Hivep3", "Cd6",
                    "Socs1", "Socs3", "Tox", "Myb",
                   "Ccnd2", "Ltb", "Ccnd3", "Ikzf2", "Cish", "Pim1")

# Scaling to 98th percentile so the scale isn't biased by high outliers
scale_98 <- function(x, na.rm = TRUE) {rescale(x, to = c(0,1), from = c(min(x), quantile(x, .98)))}

denoised_annotated_scaled <- denoised_annotated %>%
  select(all_of(genes_to_plot), Lineage_by_genotypeSlingshotWT, mean_pseudotime, Genotype, Mouse) %>%
  mutate_at(genes_to_plot, scale_98) %>% # scale is standard scaling, rescale is min/max scaling to 0,1
 pivot_longer(cols = all_of(genes_to_plot), names_to = "Gene")
```

```{r Add protein data}
denoised_prot_annotated <- denoised_proteins %>%
  left_join(pseudotime_data %>% select(Barcode, mean_pseudotime, Lineage_by_genotypeSlingshotWT, Genotype, Mouse), by = "Barcode") %>%
  filter(Lineage_by_genotypeSlingshotWT %in% c("CD4", "CD8", "WT"))

proteins_to_plot <- c("ADT_CD4_A0001", "ADT_CD8a_A0002", "ADT_CD8b(Ly-3)_A0230", "ADT_CD24_A0212", "ADT_CD5_A0111", "ADT_CD69_A0197", "ADT_TCRbchain_A0120",
                      "ADT_CD55(DAF)_A0558", "ADT_integrinb7_A0214",
                      "ADT_CD127(IL-7Ra)_A0198", "ADT_CD62L_A0112")

denoised_annotated_prot_scaled <- denoised_prot_annotated %>%
  select(all_of(proteins_to_plot), Lineage_by_genotypeSlingshotWT, mean_pseudotime, Genotype, Mouse) %>%
  mutate_at(proteins_to_plot, scale_98) %>%
  pivot_longer(cols = all_of(proteins_to_plot), names_to = "Gene")

# Combine RNA and protein data for plotting
denoised_all <- rbind(denoised_annotated_prot_scaled, denoised_annotated_scaled) %>%
  mutate(Feature = str_replace(str_replace(str_replace(Gene, "ADT_", ""), "_A[0-9]*", ""), "[(].*", ""))
```

```{r Plotting settings}
redcol <- brewer.pal(3, "Set1")[1]
bluecol <- brewer.pal(3, "Set1")[2]
greencol <- brewer.pal(3, "Set1")[3]
```

## Plot TCR target genes over pseudotime

```{r TCR targets}
tcr_targets <- factor(c("Cd5", "Dusp5", "Gata3", "Cd69", "Cd28", "Egr1", "Lef1", "Nr4a1"), levels = c("Cd5", "Dusp5", "Gata3", "Cd69", "Cd28", "Egr1", "Lef1", "Nr4a1"))

denoised_all_arranged_tcr <- denoised_all %>% 
  filter(Feature %in% tcr_targets, Lineage_by_genotypeSlingshotWT %in% c("CD4", "CD8")) %>% 
  mutate(Feature = factor(Feature, levels = levels(tcr_targets))) %>%
  arrange(factor(Feature, levels = tcr_targets))

ggplot(denoised_all_arranged_tcr, aes(x = mean_pseudotime, y = value, color = Lineage_by_genotypeSlingshotWT)) +
  geom_smooth(se = FALSE, method = "loess", span = 0.5) +
  scale_color_manual(values = c(redcol, bluecol)) +
  facet_wrap("Feature", scales = "free_y", ncol = 2) +
  theme_classic() +
  labs(x = "Pseudotime", color = "Lineage", y = "Scaled denoised RNA expression", title = "TCR targets")

ggsave("/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/DE_thymus/figures/tcr_targets.pdf", bg = "transparent", height = 5, width = 2)
```

## Plot calibrated feature examples

```{r Plot calibrated feature examples}
calibrated_features <- denoised_all %>% 
  mutate(Feature = str_replace(str_replace(Gene, "ADT_", ""), "_A[0-9]*", "")) %>% # include CD55(DAF)
  filter(Feature %in% c("Rag1", "Cxcr4", "Trbc1", "Ccr9", "Ccr4", "Ccr7", "H2-K1", "Klf2", "S1pr1", "CD24", "CD62L", "CD55(DAF)"), Lineage_by_genotypeSlingshotWT %in% c("CD4", "CD8")) 
calibrated_features$Feature <- factor(calibrated_features$Feature, levels = c("Rag1", "Cxcr4", "Trbc1", "Ccr9", "Ccr4", "Ccr7", "H2-K1", "Klf2", "S1pr1", "CD24", "CD62L", "CD55(DAF)"))

ggplot(calibrated_features, aes(x = mean_pseudotime, y = value, color = Lineage_by_genotypeSlingshotWT)) + 
  geom_smooth(se = FALSE, method = "loess", span = 0.5) +
  scale_color_manual(values = c(redcol, bluecol)) +
  facet_wrap("Feature", scales = "free_y", ncol = 3) +
  theme_classic() +
  labs(x = "Pseudotime", color = "Lineage", y = "Scaled denoised expression")

ggsave("figures/calibrated_features.pdf", bg = "transparent", height = 4.5, width = 4.5)
```

## Plot coreceptors, transcription factors, and TCR signaling molecules over pseudotime

```{r Plot coreceptors, TFs, and TCR response}
# Add facet labels
lineage_labels <- c("MHCII-specific", "MHCI-specific")
names(lineage_labels) <- c("CD4", "CD8")

# Coreceptors
ggplot() +
  geom_smooth(denoised_all %>% filter(Feature %in% c("CD4", "CD8a", "Cd4", "Cd8a"),
                                            Lineage_by_genotypeSlingshotWT %in% c("CD4", "CD8")), 
       mapping = aes(x = mean_pseudotime, y = value, color = Feature, linetype = Feature), se = FALSE, method = "loess", span = 0.5, size = .8) +
  facet_wrap("Lineage_by_genotypeSlingshotWT", ncol = 1, labeller = labeller(Lineage_by_genotypeSlingshotWT = lineage_labels)) +
  theme_classic() +
  scale_color_manual(values = c("#F8766D", "#F8766D", "#00B6EB", "#00B6EB"), labels = c("Cd4", "CD4", "Cd8a", "CD8a")) +
  scale_linetype_manual(labels = c("Cd4", "CD4", "Cd8a", "CD8a"), values = c("longdash", "solid", "longdash", "solid")) +
  scale_x_continuous(breaks = seq(0, 15, 1), minor_breaks = NULL) +
  labs(x = "Pseudotime", color = "Feature", y = "Scaled denoised expression", linetype = "Feature")
ggsave("figures/timing_coreceptors_rev1.pdf", bg = "transparent", height = 3.5, width = 5)

# Transcription factors
ggplot() +
  stat_smooth(denoised_all %>% filter(Feature %in% c("Gata3", "Zbtb7b", "Runx3"),
                                            Lineage_by_genotypeSlingshotWT %in% c("CD4", "CD8")), 
       mapping = aes(x = mean_pseudotime, y = value, color = Feature, linetype = Feature), geom = "line", se = FALSE, method = "loess", span = 0.5, size = .8) +
  facet_wrap("Lineage_by_genotypeSlingshotWT", ncol = 1, labeller = labeller(Lineage_by_genotypeSlingshotWT = lineage_labels)) +
  theme_classic() +
  scale_x_continuous(breaks = seq(0, 15, 1), minor_breaks = NULL) +
  scale_color_manual(values = c("#A58AFF", "#00C094", "#FB61D7"), labels = c("Gata3", "Runx3", "Zbtb7b")) +
  scale_linetype_manual(labels = c("Gata3", "Runx3", "Zbtb7b"), values = c("longdash", "longdash", "longdash")) +  
  labs(x = "Pseudotime", color = "Feature", y = "Scaled denoised expression", linetype = "Feature")  
ggsave("figures/timing_masterregs_coreceptors_rev1.pdf", bg = "transparent", height = 3.5, width = 5)

# TCR response
ggplot() +
  stat_smooth(denoised_all %>% filter(Feature %in% c("Cd69", "Egr1"),
                                            Lineage_by_genotypeSlingshotWT %in% c("CD4", "CD8")), 
       mapping = aes(x = mean_pseudotime, y = value, color = Feature, linetype = Feature), geom = "line", se = FALSE, method = "loess", span = 0.5, size = .8) +
  facet_wrap("Lineage_by_genotypeSlingshotWT", ncol = 1, labeller = labeller(Lineage_by_genotypeSlingshotWT = lineage_labels)) +
  theme_classic() +
  scale_x_continuous(breaks = seq(0, 15, 1), minor_breaks = NULL) +
  scale_color_manual(values = c("Cd69" = "#C49A00", "Egr1" = "#53B400"),
                     breaks = c("Cd69", "Egr1")) + 
  scale_linetype_manual(values = c("Cd69" = "longdash", "Egr1" = "longdash"),
                        breaks = c("Cd69", "Egr1")) + 
  labs(x = "Pseudotime", color = "Feature", y = "Scaled denoised expression", linetype = "Feature")  
ggsave("figures/timing_tcrresponse_masterregs_coreceptors_rev1.pdf", bg = "transparent", height = 3.5, width = 5)
```
## Plot RNA vs protein example (CD69)

```{r CD69}
# Pseudotime plot for CD69 RNA/protein to highlight RNA changes precede protein

# Add facet labels
lineage_labels <- c("MHCII-specific", "MHCI-specific")
names(lineage_labels) <- c("CD4", "CD8")

ggplot() +
  geom_smooth(denoised_all %>% filter(Feature %in% c("Cd69", "CD69"),
                                            Lineage_by_genotypeSlingshotWT %in% c("CD4", "CD8")), 
       mapping = aes(x = mean_pseudotime, y = value, linetype = Feature), se = FALSE, method = "loess", span = 0.5, size = .8, color = "#C49A00") +  
  facet_wrap("Lineage_by_genotypeSlingshotWT", ncol = 1, labeller = labeller(Lineage_by_genotypeSlingshotWT = lineage_labels)) +
  theme_classic() +
  scale_color_manual(values = c("#C49A00", "#C49A00"), labels = c("Cd69 RNA", "CD69 protein")) +
  scale_linetype_manual(labels = c("Cd69 RNA", "CD69 Protein"), values = c("longdash", "solid")) +
  scale_x_continuous(breaks = seq(0, 15, 1), minor_breaks = NULL) +
  labs(x = "Pseudotime", color = "Feature", y = "Scaled denoised expression", linetype = "Feature")

ggsave("figures/timing_CD69_RNA_protein.pdf", bg = "transparent", height = 3.5, width = 5)
```

## Plot features over pseudotime
```{r By genotype}
markers_by_genotype <- factor(c("Cd4", "Cd8a", "Cd8b1", "Cd5", "Cd69", "Il7r",
                         "CD4", "CD8a", "CD8b", "CD5", "CD69", "CD127",
                         "Egr1", "Egr2", "Id2", "Id3", "Tcf12", "Tcf3",
                         "Gata3", "Zbtb7b", "Runx3", "Dusp2", "Dusp5", "Lef1"))

denoised_all_arranged <- denoised_all %>% 
  filter(Feature %in% markers_by_genotype, Lineage_by_genotypeSlingshotWT %in% c("CD4", "CD8")) %>% 
  arrange(factor(Feature, levels = markers_by_genotype)) %>%
  mutate(Feature = factor(Feature, levels = markers_by_genotype)) %>%
  mutate(Genotype = factor(recode(Genotype, B2M = "MHCI-/-", MHC2 = "MHCII-/-", OT1 = "OT-I", "OT2" = "OT-II"), levels = c("B6", "AND", "OT-II", "MHCI-/-", "F5", "OT-I", "MHCII-/-")))

ggplot(denoised_all_arranged,
       aes(x = mean_pseudotime, y = value, color = Genotype)) +
  geom_smooth(se = FALSE, method = "loess", span = 0.5, size = .6) +
  scale_color_manual(values = c("#fddbc7", "#ef8a62", "#b2182b", "#d1e5f0", "#67a9cf", "#2166ac")) + 
  facet_wrap("Feature", scales = "free_y", ncol = 6) +
  theme_classic() +
  labs(x = "Pseudotime", color = "Genotype", y = "Scaled denoised expression")

ggsave("figures/markers_by_genotype_pseudotime_rev1.pdf", bg = "transparent", height = 5, width = 10)
```

# In-silico FACS
## Create gates

```{r Log protein counts for FACS}
denoised_proteins_facs <- denoised_proteins

# Shorten protein names for plotting
short_names <- str_replace(str_replace(names(denoised_proteins_facs[, 1:dim(denoised_proteins_facs)[2]-1]), "ADT_", ""), "_A[0-9]*", "")
colnames(denoised_proteins_facs) <- c(short_names, "Barcode")

denoised_proteins_facs[, 1:dim(denoised_proteins_facs)[2]-1] <- log1p(denoised_proteins_facs[, 1:dim(denoised_proteins_facs)[2]-1])   # take log of all proteins 
denoised_proteins_facs <- denoised_proteins_facs %>%
  left_join(pseudotime_data %>% select(Barcode, mean_pseudotime, Pseudotime_bin, Lineage_by_genotypeSlingshotWT, Genotype, Mouse, UMAP1, UMAP2), by = "Barcode") %>% 
  filter(Lineage_by_genotypeSlingshotWT %in% c("CD4", "CD8", "WT"))

denoised_proteins_facs <- rename(denoised_proteins_facs, "Lineage" = Lineage_by_genotypeSlingshotWT)
denoised_proteins_facs$Lineage <- factor(denoised_proteins_facs$Lineage, levels = c("WT", "CD4", "CD8"))
```

```{r Prepare RNA data}
denoised_rna_facs <- denoised_genes %>%
  select(Gata3, Runx3, Zbtb7b, Il7r, Egr1, Cd69, Cd4, Cd8a, Barcode)

denoised_facs <- denoised_proteins_facs %>%
  left_join(denoised_rna_facs, by = "Barcode")

denoised_facs <- denoised_facs %>% # add log of RNA values
  mutate(Gata3_log = log(Gata3)) %>%
  mutate(Runx3_log = log(Runx3)) %>%
  mutate(Zbtb7b_log = log(Zbtb7b)) %>%
  mutate(Il7r_log = log(Il7r)) %>%
  mutate(Cd69_log = log(Cd69)) %>%
  mutate(Egr1_log = log(Egr1)) 
```

```{r Make gates}
denoised_facs <- denoised_facs %>%
  mutate(CD24_gate = case_when(CD24 > 2.5 ~ "CD24+",
                               CD24 < 2.5 ~ "CD24-"))

denoised_facs <- denoised_facs %>%
  mutate(IL7Ra_gate = case_when(`CD127(IL-7Ra)` < 0.9 ~ "CD127-",
                               `CD127(IL-7Ra)` > 0.9 ~ "CD127+"))

denoised_facs <- denoised_facs %>%
  mutate(CD69_gate = case_when(CD69 < 2.55 ~ "CD69-",
                               CD69 > 2.55 ~ "CD69+"))

denoised_facs <- denoised_facs %>%
  mutate(CD4_CD8_gate = case_when(CD4 > 4.4 & CD8a > 4.4 ~ "DP",
                                  CD4 > 4.4 & CD8a < 4.4 & CD8a > 2.6 ~ "CD4+CD8low",
                                  CD8a > 4.4 & CD4 < 4.4 & CD4 > 2.6 ~ "CD4lowCD8+",
                                  CD4 > 4.4 & CD8a < 2.6 ~ "CD4+",
                                  CD8a > 4.4 & CD4 < 2.6 ~ "CD8+"))
```

```{r Combine gates}
denoised_facs <- denoised_facs %>%
  mutate(CD127_gate = IL7Ra_gate)

denoised_facs_combined <- denoised_facs %>%
  unite("combined_gate", c("CD4_CD8_gate", "CD69_gate", "CD127_gate"), remove = FALSE)

denoised_facs_combined <- denoised_facs_combined %>%
  unite("CD127_CD69_gate", c("CD127_gate", "CD69_gate"), remove = FALSE)
```

```{r Add TCRb gate}
denoised_facs_combined <- denoised_facs_combined %>%
  mutate(TCRbchain_gate = case_when(TCRbchain > 3.8 ~ "TCRbhi",
                                    TCRbchain < 3.8 ~ "TCRblo"))
```

```{r Alternative gate without CD127}
denoised_facs_combined <- denoised_facs_combined %>%
  unite("combined_CD4_CD8_CD69", c("CD4_CD8_gate", "CD69_gate"), remove = FALSE)
denoised_facs_combined$Lineage <- factor(denoised_facs_combined$Lineage, levels = c("WT", "CD4", "CD8"))
```

```{r New gates v4}
denoised_facs_combined <- denoised_facs_combined %>%
  mutate(summary_gates_v4 = case_when(combined_gate %in% c("DP_CD69-_CD127-") ~ "DP1", 
                                   ((combined_gate %in% c("DP_CD69+_CD127-"))) ~ "DP2", 
                                   combined_CD4_CD8_CD69 %in% c("CD4+CD8low_CD69+") ~ "CD4+CD8low", 
                                   ((combined_gate %in% c("DP_CD69+_CD127+")) & (TCRbchain_gate == "TCRbhi")) ~ "DP3", 
                                   combined_gate %in% c("CD8+_CD69+_CD127+", "CD4lowCD8+_CD69+_CD127+") ~ "SemimatureCD8", 
                                   combined_gate %in% c("CD4+_CD69+_CD127+") ~ "SemimatureCD4",
                                   combined_gate %in% c("CD8+_CD69-_CD127+", "CD4lowCD8+_CD69-_CD127+") ~ "MatureCD8", 
                                   combined_gate %in% c("CD4+_CD69-_CD127+") ~ "MatureCD4" )) 

denoised_facs_combined$summary_gates_v4 <- factor(denoised_facs_combined$summary_gates_v4,
                                               levels = c("DP1", "DP2", "CD4+CD8low", "DP3", "SemimatureCD4", "SemimatureCD8", "MatureCD4", "MatureCD8"))
```

```{r Plot histogram of gates over pseudotime}
ggplot(denoised_facs_combined %>% filter(!(is.na(summary_gates_v4))), aes(x = mean_pseudotime, fill = summary_gates_v4)) +
  geom_histogram(bins = 30) +
  labs(x = "Pseudotime", y = "Cells", fill = "Gate") +
  scale_fill_manual(values = gate_colors) +
  xlim(-.1, 16.1) +
  facet_grid(rows= vars(summary_gates_v4), cols = vars(Lineage), switch = "y")+
  theme_light() +
  theme(strip.text.y.left = element_text(angle = 0))

ggsave("figures/histogram_gated_facet_lineage.png", dpi = 1000, bg = "transparent", height = 4.5, width = 7)
```

## Schematic timeline

```{r Find thresholds for gates v4}
# Pseudotime thresholds between populations for schematic timeline

# DP1 to DP2 in all cells
threshold_data <- denoised_facs_combined %>%
  select(summary_gates_v4, mean_pseudotime, Lineage) %>%
  filter(summary_gates_v4 %in% c("DP1", "DP2")) %>%
  select(response = summary_gates_v4, predictor = mean_pseudotime)

roc1 <- roc(threshold_data, response = response, predictor = predictor)
thresh1 <- coords(roc1, "best", ret = "threshold", best.method = "youden")[[1]]

# DP2 to CD4+CD8low in all cells
threshold_data <- denoised_facs_combined %>%
  select(summary_gates_v4, mean_pseudotime, Lineage) %>%
  filter(summary_gates_v4 %in% c("DP2", "CD4+CD8low")) %>%
  select(response = summary_gates_v4, predictor = mean_pseudotime)

roc2 <- roc(threshold_data, response = response, predictor = predictor, levels = c("DP2", "CD4+CD8low"))
thresh2 <- coords(roc2, "best", ret = "threshold", best.method = "youden")[[1]]

# Threshold leaving CD4+CD8low is nearly identical between the two lineages (9.3), so draw a single threshold instead of two
threshold_data <- denoised_facs_combined %>%
  select(summary_gates_v4, mean_pseudotime, Lineage) %>%
  filter(summary_gates_v4 %in% c("CD4+CD8low", "SemimatureCD4", "DP3")) %>%
  mutate(new_gate = case_when(summary_gates_v4 == "CD4+CD8low" ~ "CD4+CD8low",
                              summary_gates_v4 %in% c("SemimatureCD4", "DP3") ~ "rest")) %>%
  select(response = new_gate, predictor = mean_pseudotime)

roc3 <- roc(threshold_data, response = response, predictor = predictor, levels = c("CD4+CD8low", "rest"))
thresh3 <- coords(roc3, "best", ret = "threshold", best.method = "youden")[[1]]

# DP3 to SemimatureCD8
threshold_data <- denoised_facs_combined %>%
  select(summary_gates_v4, mean_pseudotime, Lineage) %>%
  filter(summary_gates_v4 %in% c("DP3", "SemimatureCD8")) %>%
  select(response = summary_gates_v4, predictor = mean_pseudotime)

roc5 <- roc(threshold_data, response = response, predictor = predictor, levels = c("DP3", "SemimatureCD8"))
thresh5 <- coords(roc5, "best", ret = "threshold", best.method = "youden")[[1]]

# Semimature to Mature
threshold_data <- denoised_facs_combined %>%
  select(summary_gates_v4, mean_pseudotime, Lineage) %>%
  filter(summary_gates_v4 %in% c("SemimatureCD4", "MatureCD4", "SemimatureCD8", "MatureCD8")) %>%
  mutate(new_gate = case_when(summary_gates_v4 %in% c("SemimatureCD4", "SemimatureCD8") ~ "Semimature",
                              summary_gates_v4 %in% c("MatureCD4", "MatureCD8") ~ "Mature")) %>%
  select(response = new_gate, predictor = mean_pseudotime)

roc4 <- roc(threshold_data, response = response, predictor = predictor, levels = c("Semimature", "Mature"))
thresh4 <- coords(roc4, "best", ret = "threshold", best.method = "youden")[[1]]
```

```{r Plot timeline histograms}
lineage_labels <- c("MHCII-specific", "MHCI-specific")
names(lineage_labels) <- c("CD4", "CD8")

gate_colors <- c("#D783FF", "#942193", "#FF8AD8", "#7A81FF",
                 "#FF2F92", "#0096FF",
                 "#941751", "#005493")
names(gate_colors) <- c("DP1", "DP2", "CD4+CD8low", "DP3", "SemimatureCD4", "SemimatureCD8", "MatureCD4", "MatureCD8")

# CD8
ggplot(denoised_facs_combined %>% filter(!(is.na(summary_gates_v4)), Lineage %in% c("CD8"), summary_gates_v4 %in% c("DP1", "DP2", "CD4+CD8low", "DP3", "SemimatureCD8", "MatureCD8")), aes(x = mean_pseudotime, after_stat(density), fill = summary_gates_v4)) +
  geom_histogram(bins = 50, alpha = 0.5, position = "identity") +
  labs(x = "Pseudotime", y = "Density", fill = "Gate") +
  scale_fill_manual(values = gate_colors[c(1,2,3,4,6,8)]) +
  theme_classic() +
  facet_wrap(vars(Lineage), ncol = 3, labeller = labeller(Lineage = lineage_labels)) +
  geom_vline(xintercept = thresh1) +
  geom_vline(xintercept = thresh2) +
  geom_vline(xintercept = thresh3) +
  geom_vline(xintercept = thresh4) +
  geom_vline(xintercept = thresh5) +
  theme(legend.position = "none") +
  xlim(-0.2,16.3)
ggsave("figures/timeline_CD8gates.png", dpi = 1000, bg = "transparent", height = 1.5, width = 7.5)

# CD4
ggplot(denoised_facs_combined %>% filter(!(is.na(summary_gates_v4)), Lineage %in% c("CD4"), summary_gates_v4 %in% c("DP1", "DP2", "CD4+CD8low", "SemimatureCD4", "MatureCD4")), aes(x = mean_pseudotime, after_stat(density), fill = summary_gates_v4)) +
  geom_histogram(bins = 50, alpha = 0.5, position = "identity") +
  labs(x = "Pseudotime", y = "Density", fill = "Gate") +
  theme_classic() +
  scale_fill_manual(values = gate_colors[c(1,2,3,5,7)]) +
  facet_wrap(vars(Lineage), ncol = 3, labeller = labeller(Lineage = lineage_labels)) +
  geom_vline(xintercept = thresh1) +
  geom_vline(xintercept = thresh2) +
  geom_vline(xintercept = thresh3) +
  geom_vline(xintercept = thresh4) +
  theme(legend.position = "none") +
  xlim(-0.2,16.3)
ggsave("figures/timeline_CD4gates.png", dpi = 1000, bg = "transparent", height = 1.5, width = 7.5)
```

```{r Use thresholds to make schematic gate timeline large}
schematic_timeline <- tibble(Pseudotime = c(thresh1, thresh2 - thresh1, thresh3 - thresh2, thresh4-thresh3, 16-thresh4, thresh1, thresh2-thresh1, thresh3-thresh2, thresh5-thresh3, thresh4-thresh5, 16-thresh4), 
                             Gate = factor(c("DP1", "DP2", "CD4+CD8low", "SemimatureCD4", "MatureCD4", "DP1", "DP2", "CD4+CD8low", "DP3", "SemimatureCD8", "MatureCD8"), levels = c("DP1", "DP2", "CD4+CD8low", "DP3", "SemimatureCD4", "SemimatureCD8", "MatureCD4", "MatureCD8")), 
                             Lineage = c(rep("CD4", 5), rep("CD8", 6)))
```

```{r Skinny schematic}
schematic_timeline$Lineage <- factor(schematic_timeline$Lineage, levels = c("CD8", "CD4"))
schematic_timeline <- schematic_timeline %>%
  mutate(Gate_short = recode(Gate, SemimatureCD4 = "SM CD4",
                             SemimatureCD8 = "SM CD8"))

# With text
ggplot(schematic_timeline, aes(fill = Gate, x = Pseudotime, y = Lineage)) +
  geom_bar(position = position_stack(reverse = TRUE), stat = "identity", width = .9, color = "white") +
  theme_classic() +
  geom_text(aes(label = Gate_short), position = position_stack(reverse = TRUE, vjust = .5), angle = 0) +
  scale_fill_manual(values = gate_colors) +
  theme(legend.position = "none", aspect.ratio = 1/10)
ggsave("figures/timeline_schematic.png", dpi = 1000, bg = "transparent", height = 2, width = 7)
```

```{r Schematic timeline no text}
ggplot(schematic_timeline, aes(fill = Gate, x = Pseudotime, y = Lineage)) +
  geom_bar(position = position_stack(reverse = TRUE), stat = "identity", width = .9, color = "white") +
  theme_classic() +
  scale_fill_manual(values = gate_colors) +
  guides(fill = guide_legend(ncol = 2)) +
  theme(legend.position = "right", aspect.ratio = 1/10)
ggsave("figures/timeline_schematic_notext.pdf", dpi = 1000, bg = "transparent", height = 3.5, width = 7)
```

## FACS plots using gates
### Zbtb7b vs Runx3

```{r Zbtb7b vs Runx3 by lineage colored by time}
ggplot(denoised_facs_combined, aes(Runx3_log, Zbtb7b_log, color = mean_pseudotime)) +
  geom_point(size = .6, stroke = 0) +
  theme_light() +
  scale_color_gradientn(colors = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)) +
  labs(color = "Pseudotime", x = "log(Runx3)", y = "log(Zbtb7b)") +
  facet_wrap(vars(Lineage), nrow = 1) +
  theme(strip.text.y.left = element_text(angle = 0)) +
  coord_equal() +
  geom_vline(xintercept = (-8.3))+
  geom_hline(yintercept = (-10))

ggsave("figures/FACS_Runx3_Zbtb7b_facet_lineage_colorpseudotime.png", dpi = 1000, bg = "transparent", height = 3, width = 5)
```

```{r Zbtb7b vs Runx3 by time bins}
denoised_facs_combined <- denoised_facs_combined %>%
  mutate(Bin_ranges = recode(Pseudotime_bin, `0` = "0-2",
                             `1` = "2-4",
                             `2` = "4-6",
                             `3` = "6-8",
                             `4` = "8-10",
                             `5` = "10-12",
                             `6` = "12-14",
                             `7` = "14-16"))

ggplot(denoised_facs_combined %>% filter(!(is.na(summary_gates_v4))), aes(Runx3_log, Zbtb7b_log, color = mean_pseudotime)) +
  geom_point(size = .5, stroke = 0) +
  theme_light() +
  scale_color_gradientn(colors = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)) +
  labs(color = "Pseudotime", x = "log(Runx3)", y = "log(Zbtb7b)") +
  facet_grid(cols = vars(Bin_ranges), rows = vars(Lineage)) + 
  geom_vline(xintercept = (-8.3))+
  geom_hline(yintercept = (-10)) +
  coord_equal()

ggsave("figures/FACS_Runx3_Zbtb7b_facet_time_lineage_colorpseudotime.png", dpi = 1000, bg = "transparent", height = 4.5, width = 7.5)
ggsave("figures/FACS_Runx3_Zbtb7b_facet_time_lineage_colorpseudotime.pdf", height = 4.5, width = 7.5, bg = "transparent")
```

### CD4 vs CD8

```{r All cells FACS CD4 CD8}
ggplot(denoised_facs_combined, aes(CD8a, CD4, color = mean_pseudotime)) +
  geom_point(size = 1, stroke = 0) +
  theme_classic() +
  scale_color_gradientn(colors = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)) +
  labs(x = "CD8a", y = "CD4", color = "Pseudotime") +
  coord_equal() +
  geom_vline(xintercept = 4.4, color = "black") + 
  geom_hline(yintercept = 4.4, color = "black") +
  geom_segment(x = 2.6, y = 4.4, xend = 2.6, yend = 7, color = "black", size = .3) +
  geom_segment(x = 4.4, y = 2.6, xend = 7, yend = 2.6, color = "black", size = .3, linetype = "dashed") +
  geom_density_2d(aes(CD8a, CD4), color = "black", bins = 20, size = .3)

ggsave("figures/FACS_CD4_CD8_time.png", dpi = 1000, bg = "transparent", height = 4.5, width = 4.5)
```

```{r Facet by lineage CD4 CD8}
ggplot(denoised_facs_combined, aes(CD8a, CD4, color = mean_pseudotime)) +
  geom_point(size = .5, stroke = 0) +
  theme_classic() +
  scale_color_gradientn(colors = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)) +
  labs(x = "CD8a", y = "CD4", color = "Pseudotime") +
  facet_wrap("Lineage", ncol = 1) +
  coord_equal() +
  geom_vline(xintercept = 4.4, color = "black") + 
  geom_hline(yintercept = 4.4, color = "black") +
  geom_segment(x = 2.6, y = 4.4, xend = 2.6, yend = 7, color = "black", size = .2) +
  geom_segment(x = 4.4, y = 2.6, xend = 7, yend = 2.6, color = "black", size = .2, linetype = "dashed")

ggsave("figures/FACS_CD4_CD8_time_facet_lineage.png", dpi = 1000, bg = "transparent", height = 3.5, width = 3.5)
```
```{r Facet by time and lineage CD4 CD8}
ggplot(denoised_facs_combined, aes(CD8a, CD4, color = mean_pseudotime)) +
  geom_point(size = .5, stroke = 0) + 
  theme_light() +
  scale_color_gradientn(colors = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)) +
  labs(color = "Pseudotime") +
  facet_grid(cols = vars(Bin_ranges), rows = vars(Lineage)) +
  coord_equal() +
  geom_vline(xintercept = 4.4, color = "black") + 
  geom_hline(yintercept = 4.4, color = "black") +
  geom_segment(x = 2.6, y = 4.4, xend = 2.6, yend = 7, color = "black", size = .2) +
  geom_segment(x = 4.4, y = 2.6, xend = 7, yend = 2.6, color = "black", size = .2, linetype = "dashed")

ggsave("figures/FACS_CD4_CD8_facet_time_lineage.png", dpi = 1000, height = 4.5, width = 7.5, bg = "transparent")
ggsave("figures/FACS_CD4_CD8_facet_time_lineage.pdf", height = 4.5, width = 7.5, bg = "transparent")
```

### CD69 vs CD127
```{r Plot all cells CD69 CD127}
ggplot(denoised_facs_combined, aes(`CD127(IL-7Ra)`, CD69, color = mean_pseudotime)) + 
  geom_point(size = 1, stroke = 0) +
  theme_classic() +
  scale_color_gradientn(colors = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)) +
  labs(color = "Pseudotime") +
  coord_equal() +
  geom_hline(yintercept = 2.55) +
  geom_vline(xintercept = 0.9) +
  geom_density_2d(color = "black", bins = 20, size = 0.3)

ggsave("figures/FACS_CD69_CD127_time.png", dpi = 1000, bg = "transparent", height = 4.5, width = 4.5)
```
```{r Facet by lineage CD69 CD127}
ggplot(denoised_facs_combined, aes(`CD127(IL-7Ra)`, CD69, color = mean_pseudotime)) +
  geom_point(size = .5, stroke = 0) +
  theme_classic() +
  scale_color_gradientn(colors = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)) +
  labs(color = "Pseudotime") +
  facet_wrap("Lineage", ncol = 1) +
  coord_equal() +
  geom_hline(yintercept = 2.55) +
  geom_vline(xintercept = 0.9)

ggsave("figures/FACS_CD69_CD127_time_facet_lineage.png", dpi = 1000, bg = "transparent", height = 4, width = 4)
```

```{r CD69 CD127 by time and lineage}
ggplot(denoised_facs_combined, aes(`CD127(IL-7Ra)`, CD69, color = mean_pseudotime)) + 
  geom_point(size = .5, stroke = 0) + 
  theme_light() +
  scale_color_gradientn(colors = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)) +
  labs(color = "Pseudotime") +
  facet_grid(cols = vars(Bin_ranges), rows = vars(Lineage)) +
  coord_equal() +
  geom_hline(yintercept = 2.55) +
  geom_vline(xintercept = 0.9)

ggsave("figures/FACS_CD69_CD127_facet_time_lineage.png", dpi = 1000, height = 4.5, width = 7.5, bg = "transparent")
ggsave("figures/FACS_CD69_CD127_facet_time_lineage.pdf", height = 4.5, width = 7.5, bg = "transparent")
```

### CD5 vs TCRb
```{r Plot DP cells CD5 TCRb}
ggplot(denoised_facs_combined %>% filter(CD4_CD8_gate == "DP"), aes(`TCRbchain`, CD5, color = mean_pseudotime)) + 
  geom_point(size = 1, stroke = 0) +
  theme_classic() +
  scale_color_gradientn(colors = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)) +
  labs(color = "Pseudotime") +
  coord_equal() +
  geom_density_2d(color = "black", bins = 20, size = 0.3) +
  geom_vline(xintercept = 3.8)

ggsave("figures/FACS_TCRb_CD5_time.png", dpi = 1000, bg = "transparent", height = 4.5, width = 4.5)
```
```{r DP cells plot by lineage CD5 TCRb}
ggplot(denoised_facs_combined%>% filter(CD4_CD8_gate == "DP"), aes(`TCRbchain`, CD5, color = mean_pseudotime)) +
  geom_point(size = .5, stroke = 0) +
  theme_classic() +
  scale_color_gradientn(colors = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)) +
  labs(color = "Pseudotime") +
  facet_wrap("Lineage", ncol = 1) +
  coord_equal() +
  geom_vline(xintercept = 3.8)

ggsave("figures/FACS_TCRb_CD5_time_facet_lineage.png", dpi = 1000, bg = "transparent", height = 3.5, width = 3.5)
```

## Gates on UMAP

```{r Plot CITE-seq gates on UMAP}
gate_colors <- c("#D783FF", "#942193", "#FF8AD8", "#7A81FF",
                 "#FF2F92", "#0096FF",
                 "#941751", "#005493")
names(gate_colors) <- c("DP1", "DP2", "CD4+CD8low", "DP3", "SemimatureCD4", "SemimatureCD8", "MatureCD4", "MatureCD8")

ggplot(denoised_facs_combined %>% filter(!(is.na(summary_gates_v4))), mapping = aes(x = UMAP1, y = UMAP2, color = summary_gates_v4)) +
  theme_classic() +
  scale_color_manual(values = gate_colors) +
  geom_point(stroke  = 0, size = .3) +
  coord_fixed(ratio = 1) +
  labs(color = "Gate") +
  guides(colour = guide_legend(override.aes = list(size=2)))
ggsave("figures/umap_gated.png", dpi = 1000, bg = "transparent", height = 3, width = 6)
```



## Save FACS gated data

```{r Write denoised FACS data}
# write_csv(denoised_facs_combined, "gated_denoised_data_20210330.csv")
# denoised_facs_combined <- read_csv("gated_denoised_data_20210330.csv")
```

# Plot pseudotime on UMAP with Slingshot results

```{r Cells per lineage by genotype}
# Tabulate cells per lineage (including wrong lineage cells) by genotype
wronglin_pct <- table(pseudotime_data$Genotype, pseudotime_data$Lineage_by_genotypeSlingshot, useNA = "ifany")
colnames(wronglin_pct)[3] <- "Wrong_Lineage"
names(dimnames(wronglin_pct)) <- c("Genotype", "Lineage")

wronglin_pct <- as_tibble(wronglin_pct) %>%
  pivot_wider(names_from = "Lineage", values_from = n) %>%
  rowwise() %>%
  mutate(Wrong_pct = Wrong_Lineage/(CD4 + CD8 + Wrong_Lineage))
wronglin_pct # Note WT cells are unassigned, so they are included in the "wrong lineage" column here
write_csv(wronglin_pct, "Lineage_by_genotype.csv")
```


```{r Load Slingshot results}
# Needed to plot curves in pseudotime
crv <- readRDS("slingshot_20200913_removeCD8outlier.rds")

slingshot_curve1 <- as_tibble(slingCurves(crv)[[1]]$s[slingCurves(crv)[[1]]$ord, ])
slingshot_curve2 <- as_tibble(slingCurves(crv)[[2]]$s[slingCurves(crv)[[2]]$ord, ])
```

```{r Plot pseudotime}
ggplot() +
  theme_classic() +
  geom_point(time_bins, mapping = aes(x = UMAP1, y = UMAP2, color = mean_pseudotime), stroke  = 0, size = .2) +
  scale_color_gradientn(colors = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)) +
  coord_fixed(ratio = 1) +
  geom_path(slingshot_curve1, mapping = aes(x = UMAP1, y = UMAP2), size = 0.3) +
  geom_path(slingshot_curve2, mapping = aes(x = UMAP1, y = UMAP2), size = 0.3) +
  labs(color = "Pseudotime") +
  theme(legend.position = "bottom")

ggsave("figures/umap_pseudotime_curves.pdf", bg = "transparent", height = 3, width = 5)
```

```{r Plot time bins on UMAP}
ggplot() +
  theme_classic() +
  geom_point(denoised_facs_combined, mapping = aes(x = UMAP1, y = UMAP2, color = Bin_ranges), stroke  = 0, size = .3) +
  scale_color_brewer(palette = "Spectral") +
  coord_fixed(ratio = 1) +
  labs(color = "Pseudotime") +
  guides(colour = guide_legend(override.aes = list(size=2))) +
  theme(legend.position = "bottom")

ggsave("figures/umap_pseudotime_binned.png", dpi = 1000, bg = "transparent", height = 4, width = 4) 
```

# Comparison of population frequencies in CITE-seq vs flow

```{r Find WT proportions in CITE-seq data per mouse}
denoised_facs_WT <- denoised_facs_combined %>%
  filter(Lineage == "WT")

# 8 gated populations from summary gates v4
CITEseq_proportions <- as_tibble(as.data.frame.matrix(table(denoised_facs_WT$summary_gates_v4, denoised_facs_WT$Mouse)),
                                 rownames = "Gate") %>%
  mutate(across(B6_r1:B6_r5, ~.x/sum(.x)))

CITEseq_proportions$Gate <- factor(CITEseq_proportions$Gate, levels = c("DP1", "DP2", "CD4+CD8low", "DP3", "SemimatureCD4", "SemimatureCD8",
                                                                        "MatureCD4", "MatureCD8"))

proportions <- CITEseq_proportions %>%
  pivot_longer(cols = B6_r1:B6_r5, names_to = "Mouse", values_to = "Frequency") %>%
  group_by(Gate) %>%
  summarize(n = n(),
            mean = mean(Frequency),
            sd = sd(Frequency)    ) %>%
  mutate(se = sd/sqrt(n)) %>%
  mutate(Data = "CITE-seq")
```

```{r Add proportions from flow}
flow_proportions <- read_csv("Flow_adultWTthymiccellproportions.csv") %>%
  select(Gate:M5, M8:M9) # exclude M6 and M7, which were missing the gate on CD5 (all others gated with TCRb vs CD5 gate for positive selection)

flow_proportions$Gate <- factor(flow_proportions$Gate, levels = c("DP1", "DP2", "CD4+CD8low", "DP3", "SemimatureCD4", "SemimatureCD8",
                                                                        "MatureCD4", "MatureCD8"))

proportions_flow <- flow_proportions %>%
  pivot_longer(cols = M1:M9, names_to = "Mouse", values_to = "Frequency") %>%
  group_by(Gate) %>%
  summarize(n = n(),
            mean = mean(Frequency),
            sd = sd(Frequency),
            ) %>%
  mutate(se = sd/sqrt(n)) %>%
  mutate(Data = "Flow cytometry")

proportions_all <- proportions %>%
  bind_rows(proportions_flow)
```

```{r Plot all proportions}
ggplot(proportions_all, aes(fill = Data)) +
  geom_col(aes(x = Gate, y = mean, fill = Data), position = position_dodge(width = 0.9)) +
  geom_errorbar(aes(x = Gate, ymin = mean - sd, ymax = mean + sd), width = 0.25, alpha = 0.5, position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("cyan3", "blueviolet")) +
  scale_x_discrete(labels=c("DP1", "DP2", "CD4+CD8low", "DP3", "CD4+SM", "CD8+SM", "CD4+Mat", "CD8+Mat")) +
  theme_classic() +
  theme(legend.position = c(0.8, 0.8), axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(y = "Frequency")

ggsave("figures/cell_type_proportions.pdf", bg = "transparent", height = 3.5*.95, width = 5*.95)
```

```{r Plot all proportions with replicates overlaid as points}
proportions_permouse_CITEseq <- CITEseq_proportions %>% pivot_longer(cols = B6_r1:B6_r5, names_to = "Mouse", values_to = "Frequency") %>% mutate(Data = "CITE-seq")
proportions_permouse_flow <- flow_proportions %>% pivot_longer(cols = M1:M9, names_to = "Mouse", values_to = "Frequency") %>% mutate(Data = "Flow cytometry")
proportions_permouse <- rbind(proportions_permouse_CITEseq, proportions_permouse_flow)

ggplot() +
  geom_col(data = proportions_all, mapping = aes(x = Gate, y = mean, fill = Data), position = position_dodge(width = 0.9)) +
  geom_point(proportions_permouse, mapping = aes(x = Gate, y = Frequency, group = Data), position = position_dodge(width = 0.9)) +
  geom_errorbar(data = proportions_all, mapping = aes(x = Gate, ymin = mean - sd, ymax = mean + sd, group = Data),
                width = 0.25, alpha = 0.5, position = position_dodge(width = 0.9)) +
  scale_fill_manual(values = c("cyan3", "blueviolet")) +
  scale_x_discrete(labels=c("DP1", "DP2", "CD4+CD8low", "DP3", "CD4+SM", "CD8+SM", "CD4+Mat", "CD8+Mat")) +
  theme_classic() +
  theme(legend.position = c(0.8, 0.8), axis.text.x = element_text(angle = 45, vjust = 1, hjust=1)) +
  labs(y = "Frequency")

ggsave("figures/cell_type_proportions.pdf", bg = "transparent", height = 3.5*.95, width = 5*.95)
```