---
title: "Slingshot on CD4/CD8 lineages in the thymus"
output: html_notebook
---

Zoë Steier

* Slingshot to find pseudotime ordering of CD4/CD8 lineages.
* Plot expression over pseudotime as heatmaps.

# Load packages

```{r Load packages}
library(tidyverse)
library(slingshot)
library(pals)
library(RColorBrewer)
library(DescTools)
library(pheatmap)

packageVersion("pals")
packageVersion("DescTools")
packageVersion("pheatmap")
```

# Load data
## Load totalVI data

```{r Load totalVI data}
totalvi_path <- "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/totalVI_thymus/thymus111_allbatches_stable_posselecting/totalVI_results/"

latent <- read_csv(gzfile(str_c(totalvi_path, "latent.csv.gz")))
umap_totalVI <- read_csv(gzfile(str_c(totalvi_path, "umap.csv.gz")))
denoised_proteins <- read_csv(gzfile(str_c(totalvi_path, "denoised_proteins.csv.gz")))
denoised_genes <- read_csv(gzfile(str_c(totalvi_path, "denoised_genes.csv.gz")))
totalVI_genes <- read_csv(gzfile(str_c(totalvi_path, "totalVI_genes.csv.gz")))
totalVI_proteins <- read_csv(gzfile(str_c(totalvi_path, "totalVI_proteins.csv.gz")))
obs <- read_csv(gzfile(str_c(totalvi_path, "obs.csv.gz")))

meta_csv <- read_csv("/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/metadata/Metadata_SeqStats_totalSeq_experiments.csv")
meta_thymus <- meta_csv %>%
  filter(Tissue == "thymus") %>%
  filter(TotalSeq_panel == "ADT111") # 19 batches in total
meta_thymus <- meta_thymus %>%
  mutate(Batch = seq(1, dim(meta_thymus)[1]) - 1) # rename batches starting at 0
```

```{r Load filtered data}
filt_path <- "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/Annotation_thymus/"
obs_filtered <- read_csv(gzfile(str_c(filt_path, "adata_positiveselection_filt_obs.csv.gz")))
filtered_cells <- obs_filtered$X1
```

## Extract totalVI data

```{r Parse totalVI data}
# Extract: clusters, umap, denoised proteins, latent, etc.
# Remove first column of dataframes - this is an index column coming from python

# totalVI umap
totalvi_umap <- as.data.frame(umap_totalVI[, -1])
colnames(totalvi_umap) <- c("UMAP1", "UMAP2")
row.names(totalvi_umap) <- obs$X1

# totalVI latent space
totalvi_latent <- as.matrix(latent[, -1])
row.names(totalvi_latent) <- obs$X1

# totalVI denoised proteins
denoised_proteins <- denoised_proteins %>%
  select(-c(X1))
colnames(denoised_proteins) <- totalVI_proteins[, 2][[1]]
denoised_proteins <- denoised_proteins %>%
  mutate(Barcode = obs$X1)

# totalVI denoised genes
denoised_genes <- denoised_genes %>%
  select(-c(X1))
colnames(denoised_genes) <- totalVI_genes[, 2][[1]]
denoised_genes <- denoised_genes %>%
  mutate(Barcode = obs$X1)
```

```{r Make metadata from totalVI}
# totalVI Leiden clusters and metadata
totalVI_meta <- obs %>%
  select(Batch = 'batch_indices', UMIs_RNA = "n_counts", UMIs_protein = "n_protein_counts",
          'n_genes', 'n_proteins', 'percent_mito', contains("leiden_totalVI_"), Barcode = "X1") %>%
  mutate_at(vars(contains("leiden_totalVI_")), as_factor) %>% # Convert cluster labels to factors
  left_join(meta_thymus %>% select(Sample, Genotype, Experiment, Replicate, Location, Batch), by = "Batch") %>% # Join with experimental metadata
  mutate(Mouse = str_c(Genotype, "_", Replicate)) %>% # Add mouse information
  mutate_at(c("Sample", "Batch", "Mouse", "Experiment", "Genotype", "Location"), as_factor) %>% # Set as factors to preserve the order of levels
  select(-"Replicate")
```

# Run Slingshot
## Prepare data for slingshot

```{r View clusters on UMAP}
set.seed(200)

# Plot clusters
totalVI_clusters <- as_tibble(cbind(totalVI_meta, as_tibble(totalvi_umap))) 

ggplot(data = totalVI_clusters, aes(x = UMAP1, y = UMAP2, color = leiden_totalVI_res0.6)) +
  geom_point(size = 0.75, stroke = 0) + 
  theme_classic() +
  scale_color_manual(values = c(brewer.paired(10))) +
  labs(color = "Cluster", title = "leiden_totalVI_res0.6") + 
  guides(colour = guide_legend(override.aes = list(size=2)))
```

```{r Filter cells by subsetting}
# Use previously filtered data to apply the above filters (additionally removing the CD8 early outliers)
totalVI_clusters_filt <- totalVI_clusters %>%
  filter(Barcode %in% filtered_cells) %>%
  mutate(slingshot_clusters = str_replace(leiden_totalVI_res0.6, "7", "1"))

ggplot(data = totalVI_clusters_filt, aes(x = UMAP1, y = UMAP2, color = slingshot_clusters)) +
  geom_point(size = .75, stroke = 0) +
  theme_classic() +
  scale_color_manual(values = c(brewer.paired(10))) +
  labs(color = "Cluster", title = "slingshot_clusters") + 
  guides(colour = guide_legend(override.aes = list(size=2)))
```

## Run slingshot

```{r Run slingshot}
# First check lineages to see if parameters need to be changed. Then run getCurves only once because it is slower.
lin <- getLineages(
  totalVI_clusters_filt[, c("UMAP1", "UMAP2")],
  totalVI_clusters_filt$slingshot_clusters,
  start.clus = "1",
  end.clus = c("0", "3")
  )
crv <- getCurves(lin, approx_points = 200)
```

```{r View lin results}
lin
```

```{r View crv results}
crv
```

```{r Plot slingshot lineages}
plot(totalVI_clusters_filt[, c("UMAP1", "UMAP2")], pch=16, cex = 0.25, col = brewer.paired(11)[as.factor(totalVI_clusters_filt$slingshot_clusters)]) 
lines(lin, lwd=2, type = 'lineages', show.constraints = TRUE, color = "black")
```

```{r Plot slingshot curves}
plot(reducedDims(crv)[,1:2], pch=16, cex = 0.25, col = brewer.paired(11)[as.factor(totalVI_clusters_filt$slingshot_clusters)]) 
lines(crv, lwd=2, col = "black")
```

```{r Save slingshot output}
getwd() #"/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/Slingshot_thymus"
saveRDS(crv, "slingshot_20200913_removeCD8outlier.rds")

# Load saved object
crv <- readRDS("slingshot_20200913_removeCD8outlier.rds")
```

## Plot slingshot results

```{r Data for plotting with ggplot}
pseudotime <- as_tibble(slingPseudotime(crv)) %>%
  mutate(difference = abs(curve1 - curve2)) 
pseudotime <- pseudotime %>%
  mutate(mean_pseudotime = rowMeans(select(pseudotime, c(curve1, curve2)), na.rm = TRUE)) %>%
  mutate(UMAP1 = reducedDims(crv)[,1]) %>%
  mutate(UMAP2 = reducedDims(crv)[,2])

# Slingshot curves for plotting
slingshot_curve1 <- as_tibble(slingCurves(crv)[[1]]$s[slingCurves(crv)[[1]]$ord, ]) 
slingshot_curve2 <- as_tibble(slingCurves(crv)[[2]]$s[slingCurves(crv)[[2]]$ord, ]) 
```

```{r Plot pseudotime ggplot}
ggplot() +
  theme_classic() +
  geom_point(pseudotime, mapping = aes(x = UMAP1, y = UMAP2, color = mean_pseudotime), stroke  = 0, size = .8) +
  scale_color_gradientn(colors = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)) + 
  coord_fixed(ratio = 1) +
  geom_path(slingshot_curve1, mapping = aes(x = UMAP1, y = UMAP2)) +
  geom_path(slingshot_curve2, mapping = aes(x = UMAP1, y = UMAP2)) +
  labs(color = "Pseudotime")
```

```{r Plot each lineage by pseudotime}
ggplot() +
  geom_point(pseudotime, mapping = aes(x = UMAP1, y = UMAP2), color = "grey75", stroke  = 0, size = .8) + 
  theme_classic() +
  geom_point(pseudotime, mapping = aes(x = UMAP1, y = UMAP2, color = curve1), stroke  = 0, size = .8) +
  scale_color_gradientn(colors = brewer.spectral(100)) +
  coord_fixed(ratio = 1) +
  labs(color = "Pseudotime")

ggplot() +
  geom_point(pseudotime, mapping = aes(x = UMAP1, y = UMAP2), color = "grey75", stroke  = 0, size = .8) +
  theme_classic() +
  geom_point(pseudotime, mapping = aes(x = UMAP1, y = UMAP2, color = curve2), stroke  = 0, size = .8) +
  scale_color_gradientn(colors = brewer.spectral(100)) +
  coord_fixed(ratio = 1) +
  labs(color = "Pseudotime")
```

```{r Plot lineage weights}
# Add lineage weights to pseudotime tibble
pseudotime <- pseudotime %>%
  mutate(weight_curve1 = slingCurveWeights(crv)[,1]) %>%
  mutate(weight_curve2 = slingCurveWeights(crv)[,2])
pseudotime_data <- as_tibble(cbind(pseudotime, totalVI_clusters_filt %>% select(-c("UMAP1", "UMAP2"))))

ggplot() +
  geom_point(pseudotime, mapping = aes(x = UMAP1, y = UMAP2), color = "grey75", stroke  = 0, size = .8) + 
  theme_classic() +
  geom_point(pseudotime, mapping = aes(x = UMAP1, y = UMAP2, color = weight_curve1), stroke  = 0, size = .8) +
  scale_color_gradientn(colors = brewer.spectral(100)) +
  coord_fixed(ratio = 1) +
  labs(color = "Lineage1 Weight")

ggplot() +
  geom_point(pseudotime, mapping = aes(x = UMAP1, y = UMAP2), color = "grey75", stroke  = 0, size = .8) + 
  theme_classic() +
  geom_point(pseudotime, mapping = aes(x = UMAP1, y = UMAP2, color = weight_curve2), stroke  = 0, size = .8) +
  scale_color_gradientn(colors = brewer.spectral(100)) +
  coord_fixed(ratio = 1) +
  labs(color = "Lineage2 Weight")

# Lineage 1 is CD4, Lineage 2 is CD8
```

## Assign cells to a lineage

```{r Assign cells to a lineage}
# First, assign cells to known lineage by genotype (WT is unknown)
# Then, exclude cells with < 0.5 weight in the intended lineage (these are likely a small percentage of cells going down the wrong lineage)
pseudotime_data_lineage <- pseudotime_data %>%
  mutate(Lineage_by_genotype = recode(as.character(Genotype), B6 = "NA", # Note that NA must be a character here for recode to work
                                      AND = "CD4",
                                      F5 = "CD8",
                                      OT1 = "CD8",
                                      OT2 = "CD4",
                                      B2M = "CD4",
                                      MHC2 = "CD8"
                                      )) %>%
  mutate(Lineage_by_genotypeSlingshot = case_when( # Excluding some cells in wrong lineage by Slingshot weights, NA for unassigned cells
    ((Lineage_by_genotype == "CD4") & (weight_curve1 > 0.5)) ~ "CD4", # CD4 when genotype is CD4, lin1 weight > 0.5
    ((Lineage_by_genotype == "CD8") & (weight_curve2 > 0.5)) ~ "CD8")) %>%  # CD8 when genotype is CD8, lin2 weight > 0.5
  mutate(LinWeight_CD4 = recode(Lineage_by_genotypeSlingshot, "CD4" = 1, "CD8" = 0)) %>% # Alternative weights: 0, 1, NA for each lineage
  mutate(LinWeight_CD8 = recode(Lineage_by_genotypeSlingshot, "CD4" = 0, "CD8" = 1))
  
pseudotime_data_lineage
```

```{r Plot new lineages}
# Select new lineage assigned cells and plot by genotype
assigned_cells <- pseudotime_data_lineage %>%
  filter(!(is.na(Lineage_by_genotypeSlingshot)))
notassigned <- pseudotime_data_lineage %>% # These are the unassigned cells (wrong lineage) that are not from B6
  filter(!(Genotype == "B6") & (is.na(Lineage_by_genotypeSlingshot)))

plot_assigned <- ggplot() +
  geom_point(assigned_cells, mapping = aes(x = UMAP1, y = UMAP2, color = Genotype), stroke  = 0, size = .7) + 
  theme_classic() +
  scale_color_manual(values = c(brewer.paired(6))) +
  coord_fixed(ratio = 1) +
  labs(color = "Lineage-assigned cells") +
  guides(colour = guide_legend(override.aes = list(size=2)))

plot_unassigned <- ggplot() +
  geom_point(assigned_cells, mapping = aes(x = UMAP1, y = UMAP2), color = "grey75", stroke  = 0, size = .7) + 
  geom_point(notassigned, mapping = aes(x = UMAP1, y = UMAP2, color = Genotype), stroke  = 0, size = .7) + 
  theme_classic() +
  scale_color_manual(values = c(brewer.paired(6))) +
  coord_fixed(ratio = 1) +
  labs(color = "Unassigned cells") +
  guides(colour = guide_legend(override.aes = list(size=2))) 

plot_assigned
plot_unassigned

# Plot colored by lineage assignment
ggplot() +
  geom_point(assigned_cells, mapping = aes(x = UMAP1, y = UMAP2, color = Lineage_by_genotypeSlingshot), stroke  = 0, size = .6) +
  theme_classic() +
  scale_color_brewer(palette = "Set1") +
  coord_fixed(ratio = 1) +
  labs(color = "Lineage assignment") +
  guides(colour = guide_legend(override.aes = list(size=2))) +
  facet_grid(rows = vars(Lineage_by_genotypeSlingshot))

ggplot() +
  geom_point(assigned_cells, mapping = aes(x = UMAP1, y = UMAP2, color = Lineage_by_genotypeSlingshot), stroke  = 0, size = .6) + 
  theme_classic() +
  scale_color_brewer(palette = "Set1") +
  coord_fixed(ratio = 1) +
  labs(color = "Lineage assignment") +
  guides(colour = guide_legend(override.aes = list(size=2)))
```

```{r Find how many cells go to correct/incorrect lineages}
cell_summary_table <- table(pseudotime_data_lineage$Genotype, pseudotime_data_lineage$Lineage_by_genotypeSlingshot, useNA = "ifany")
names(dimnames(cell_summary_table)) <- c("Genotype", "Lineage")
cell_summary <- as_tibble(cell_summary_table) %>%
  pivot_wider(names_from = "Lineage", values_from = "n") %>%
  mutate(percent_NA = `NA`/(CD4 + CD8 + `NA`)) %>%
  mutate(percent_correct =  1 - percent_NA)

table(pseudotime_data$Genotype)
cell_summary
```

```{r Save pseudotime data}
write_csv(pseudotime_data_lineage, "pseudotime_slingshot_2020913.csv")

# When reading this data, must instruct the format for the last 4 columns
pseudotime_data <- read_csv("pseudotime_slingshot_2020913.csv", col_types = cols(Lineage_by_genotype = col_character(), 
                                                                                  Lineage_by_genotypeSlingshot = col_character(),
                                                                                  LinWeight_CD4 = col_integer(),
                                                                                  LinWeight_CD8 = col_integer()))

# Bin pseudotime data
time_bins <- pseudotime_data %>%
  mutate(Pseudotime_bin = cut(mean_pseudotime, breaks = seq(0,16, by = 2), include.lowest = TRUE,
                              labels = str_c(as.character(seq(0, 7)))))
# Assign cells > 16 to the last bin
time_bins[is.na(time_bins$Pseudotime_bin), "Pseudotime_bin"] <- "7" # This is exactly 20 cells
```

# Plot pseudotime results as heatmaps
## Plot selected features in heatmap

Steps to generate pseudotime heatmaps:
1. extract data (genes x cells)
2. create small bins and average within bin (per row) with summarize_all
3. cbind CD4 and CD8
4. normalize per row (z score and winsorize to make heatmap colors visible)
5. separate CD4 and CD8
6. take rolling average over bins within lineage to smooth the rows
6. plot with constant column size so pseudotimes (early) match between lineages

```{r Load binned pseudotime data}
pseudotime_data <- read_csv("pseudotime_slingshot_2020913.csv", col_types = cols(Lineage_by_genotype = col_character(), 
                                                                                  Lineage_by_genotypeSlingshot = col_character(),
                                                                                  LinWeight_CD4 = col_integer(),
                                                                                  LinWeight_CD8 = col_integer()))

# Get binned pseudotime data
time_bins <- pseudotime_data %>%
  mutate(Pseudotime_bin = cut(mean_pseudotime, breaks = seq(0,16, by = 2), include.lowest = TRUE,
                              labels = str_c(as.character(seq(0, 7)))))
# Assign cells > 16 to the last bin
time_bins[is.na(time_bins$Pseudotime_bin), "Pseudotime_bin"] <- "7"
table(time_bins$Lineage_by_genotypeSlingshot)
```

```{r Helper information for plotting}
totalVI_proteins_parsed <- totalVI_proteins %>%
  select("totalVI_proteins" = `0`) %>%
  mutate("protein_names" = str_split(totalVI_proteins, "_", simplify = TRUE)[,2]) # simplify = TRUE returns a character matrix, then extract second column
cd4_lin_cells <- time_bins %>%
  filter(Lineage_by_genotypeSlingshot == "CD4") # 9,663 cells
cd8_lin_cells <- time_bins %>%
  filter(Lineage_by_genotypeSlingshot == "CD8") # 9,993 cells
```

```{r Pseudotime heatmap function}
# Plot heatmap for both CD4 and CD8 lineage for selected group of genes or proteins
end_time <- 16.07

plotPseudotimeHeatmap <- function(features, denoised_data, title) {
  
  ##############################
  # Select denoised data to plot
  ##############################

  # Select genes or proteins for denoised data
  if (str_detect(colnames(denoised_data)[1], "ADT")) {
    feature_type = "proteins"
    heatdata4 <- denoised_data %>%
      filter(Barcode %in% cd4_lin_cells$Barcode) %>%
      select(Barcode, (totalVI_proteins_parsed %>% filter(protein_names %in% features) %>% select(totalVI_proteins))[[1]]) %>%
      left_join(cd4_lin_cells %>% select(Barcode, mean_pseudotime, leiden_totalVI_res0.6, Pseudotime_bin), by = "Barcode")
    heatdata8 <- denoised_data %>%
      filter(Barcode %in% cd8_lin_cells$Barcode) %>%
      select(Barcode, (totalVI_proteins_parsed %>% filter(protein_names %in% features) %>% select(totalVI_proteins))[[1]]) %>%
      left_join(cd8_lin_cells %>% select(Barcode, mean_pseudotime, leiden_totalVI_res0.6, Pseudotime_bin), by = "Barcode")
    
    parsed_selected_proteins <- totalVI_proteins_parsed %>% # cleaner protein names for plotting
      filter(protein_names %in% features) %>%
      select(protein_names)
      
  } else {
    feature_type = "genes"
    heatdata4 <- denoised_data %>%
      filter(Barcode %in% cd4_lin_cells$Barcode) %>%
      select(Barcode, all_of(features)) %>%
      left_join(cd4_lin_cells %>% select(Barcode, mean_pseudotime, leiden_totalVI_res0.6, Pseudotime_bin), by = "Barcode")
    heatdata8 <- denoised_data %>%
      filter(Barcode %in% cd8_lin_cells$Barcode) %>%
      select(Barcode, all_of(features)) %>%
      left_join(cd8_lin_cells %>% select(Barcode, mean_pseudotime, leiden_totalVI_res0.6, Pseudotime_bin), by = "Barcode")
  }

  ##############################
  # Bin expression data by pseudotime and normalize
  ##############################
    
  # Bin data by pseudotime so lineages align
  bins = 500
  bindata4 <- heatdata4 %>%
    mutate(Bins = cut(mean_pseudotime, breaks = seq(0, end_time, length.out = bins), include.lowest = TRUE)) %>% # bin pseudotime
    select(-c("Barcode", "mean_pseudotime", "leiden_totalVI_res0.6", "Pseudotime_bin"))
  
  bindata8 <- heatdata8 %>%
    mutate(Bins = cut(mean_pseudotime, breaks = seq(0, end_time, length.out = bins), include.lowest = TRUE)) %>% # bin pseudotime
    select(-c("Barcode", "mean_pseudotime", "leiden_totalVI_res0.6", "Pseudotime_bin"))
  
  # Bind CD4 and CD4 so that colors are normalized together
  n_bins_4 <- dim(bindata4)[1]
  n_bins_8 <- dim(bindata8)[1]
  bindata_both_prot <- rbind(bindata4, bindata8) %>% # concatenate bins x genes for CD4, CD8
    mutate(across(where(is.numeric), scale)) %>% # standard scale for each gene
    mutate(across(where(is.numeric), ~ Winsorize(.x, probs = c(0.05, 0.95)))) # Winsorize for each gene
  
  # Separate into CD4 and CD8
  norm_CD4 <- bindata_both_prot[1:n_bins_4, ]
  norm_CD8 <- bindata_both_prot[(n_bins_4+1):(n_bins_4+n_bins_8), ]
  
  # Now summarize by bin separately for each lineage
  norm_CD4 <- norm_CD4 %>% 
    group_by(Bins) %>%
    summarize_all(mean)
  
  norm_CD8 <- norm_CD8 %>% 
    group_by(Bins) %>%
    summarize_all(mean)
  
  # Make a matrix and add row/column names
  CD4_mat <- as.matrix(norm_CD4[, 2:dim(norm_CD4)[2]])
  row.names(CD4_mat) <- norm_CD4$Bins
  if (feature_type == "proteins") {   # Rename features if they are proteins
    colnames(CD4_mat) <- parsed_selected_proteins[[1]]
  }
  CD4_mat <- t(CD4_mat)
  
  CD8_mat <- as.matrix(norm_CD8[, 2:dim(norm_CD8)[2]])
  row.names(CD8_mat) <- norm_CD8$Bins
  if (feature_type == "proteins") {   # Rename features if they are proteins
    colnames(CD8_mat) <- parsed_selected_proteins[[1]]
  }
  CD8_mat <- t(CD8_mat)

  # Matrix of data to plot
  both_prot_mat <- cbind(CD4_mat, CD8_mat)
  colnames(both_prot_mat) <- seq(1:dim(both_prot_mat)[2])
  gap_loc <- dim(CD4_mat)[2] # Where is the gap between CD4/CD8
  
  ##############################
  # Make annotations for heatmap
  ##############################
  
  # Find pseudotime values to use for annotations
  bin_ranges4 <- tibble(bin_ranges = colnames(CD4_mat)) %>%
    mutate(Pseudotime = as.numeric(str_replace(str_replace(bin_ranges, "[\\(\\[]", ""), ",.*", ""))) %>% # remove either the first ( or the first [ of the bin range
    mutate(Time_bin = cut(Pseudotime, breaks = seq(0,16, by = 2), include.lowest = TRUE, labels = str_c(as.character(seq(0, 7)))))
  bin_ranges4[is.na(bin_ranges4$Time_bin), "Time_bin"] <- "7" #   # assign cells > 16 to the last bin.
  
  bin_ranges8 <- tibble(bin_ranges = colnames(CD8_mat)) %>%
    mutate(Pseudotime = as.numeric(str_replace(str_replace(bin_ranges, "[\\(\\[]", ""), ",.*", ""))) %>%
    mutate(Time_bin = cut(Pseudotime, breaks = seq(0,16, by = 2), include.lowest = TRUE, labels = str_c(as.character(seq(0, 7)))))
  bin_ranges8[is.na(bin_ranges8$Time_bin), "Time_bin"] <- "7" 
  
  # Annotations df
  annotation_df4 <- data.frame(Pseudotime = bin_ranges4[, 2])
  row.names(annotation_df4) <- bin_ranges4$bin_ranges
  annotation_df8 <- data.frame(Pseudotime = bin_ranges8[, 2])
  row.names(annotation_df8) <- bin_ranges8$bin_ranges
  annotation_df_both <- rbind(annotation_df4, annotation_df8) # Pseudotime value
  annotation_df_both[["Lineage"]] <- c(rep("CD4", dim(annotation_df4)[1]), rep("CD8", dim(annotation_df8)[1])) # Lineage
  row.names(annotation_df_both) <- seq(1:dim(both_prot_mat)[2])
  
  # Annotation colors
  time_bin_colors <-  brewer.pal(8, "Spectral")
  names(time_bin_colors) <- as.character(seq(0, 7))
  ann_colors_both <- list(Pseudotime = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100), # the [-6] removes the lightest yellow color
                         Lineage = c("CD4" = brewer.pal(3, "Set1")[1],
                                     "CD8" = brewer.pal(3, "Set1")[2]))
  # Order of features. Arrange features by peak expression time in the CD4 lineage
  max_time4 <- apply(CD4_mat, 1, which.max)

  ##############################
  # Plot heatmap
  ##############################
  # # plot to save
  # pheatmap(both_prot_mat[order(max_time4),],
  #        cluster_cols = FALSE,
  #        cluster_rows = FALSE,
  #        border_color = NA,
  #        show_colnames = FALSE,
  #        color = colorRampPalette(viridis(10))(100),
  #        show_rownames = TRUE,
  #        annotation_col = annotation_df_both,
  #        annotation_colors = ann_colors_both,
  #        cellwidth = .5,
  #        cellheight = 10, # 10 normal, .1 for de genes, was 5 with bins
  #        treeheight_row = 0,
  #        filename = str_c("figures/", title, ".png"),
  #        gaps_col = c(gap_loc),
  #        fontsize = 8) 
  
  # plot to view
  pheatmap(both_prot_mat[order(max_time4),], 
         cluster_cols = FALSE, 
         cluster_rows = FALSE, 
         border_color = NA, 
         show_colnames = FALSE, 
         color = colorRampPalette(viridis(10))(100), 
         show_rownames = TRUE, 
         annotation_col = annotation_df_both,
         annotation_colors = ann_colors_both,
         cellwidth = .35, 
         cellheight = 5, 
         treeheight_row = 0,
         gaps_col = c(gap_loc),
         fontsize = 8)
}
```

```{r Plot maturation markers}
genes <- c("Cd4", "Cd8a", "Cd8b1", "Rag1", "Rag2", "Zbtb7b", "Runx3", "Cxcr4", "Ccr7", "Cd69", "S1pr1", "Klf2", "Gata3")
maturation_genes <- c(genes, "Trbc1", "Bcl2", "Ccr9", "Ccr4", "Cd24a", "H2-K1", "Sell", "Il7r", "Cd55", "Cd5") # markers from Hogquist et al.
plotPseudotimeHeatmap(maturation_genes, denoised_genes, "heatmap_Hogquist_markers")
```

```{r Plot marker proteins}
maturation_proteins <- c("CD4", "CD8a", "CD8b(Ly-3)", "TCRbchain", "CD24", "CD5", "CD69", "CD127(IL-7Ra)", "CD55(DAF)", "CD62L")
plotPseudotimeHeatmap(maturation_proteins, denoised_proteins, "heatmap_protein_markers")
```

## Plot all DE genes over time in heatmap

```{r Load DE features}
# Select DE features over time and plot as a heatmap
# Find genes that are DE in any time bin in either lineage

# Load DE results (small time bins)
de_features4 <- read_csv("/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/DE_thymus/DE_results/DE_CD4_overTime_smallbins.csv")
de_features8 <- read_csv("/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/DE_thymus/DE_results/DE_CD8_overTime_smallbins.csv")

# DE genes over time in both lineages
de_genes4 <- de_features4 %>%
  filter(bayes_factor > 2, lfc_median > 0.2, non_zeros_proportion1 > 0.05, !str_detect(X1, "ADT"))
de_genes8 <- de_features8 %>%
  filter(bayes_factor > 2, lfc_median > 0.2, non_zeros_proportion1 > 0.05, !str_detect(X1, "ADT"))
de_genes_both <- unique(c(de_genes4$X1, de_genes8$X1))
#length(de_genes_both) # 1113 genes
```

```{r Make DE labeled heatmap for DE genes}
denoised_data <- denoised_genes
features <- de_genes_both

heatdata4 <- denoised_data %>%
  filter(Barcode %in% cd4_lin_cells$Barcode) %>%
  select(Barcode, all_of(features)) %>%
  left_join(cd4_lin_cells %>% select(Barcode, mean_pseudotime, leiden_totalVI_res0.6, Pseudotime_bin), by = "Barcode")
heatdata8 <- denoised_data %>%
  filter(Barcode %in% cd8_lin_cells$Barcode) %>%
  select(Barcode, all_of(features)) %>%
  left_join(cd8_lin_cells %>% select(Barcode, mean_pseudotime, leiden_totalVI_res0.6, Pseudotime_bin), by = "Barcode")

bins = 500
bindata4 <- heatdata4 %>%
  mutate(Bins = cut(mean_pseudotime, breaks = seq(0, end_time, length.out = bins), include.lowest = TRUE)) %>% # bin pseudotime
  select(-c("Barcode", "mean_pseudotime", "leiden_totalVI_res0.6", "Pseudotime_bin"))
bindata8 <- heatdata8 %>%
  mutate(Bins = cut(mean_pseudotime, breaks = seq(0, end_time, length.out = bins), include.lowest = TRUE)) %>% # bin pseudotime
  select(-c("Barcode", "mean_pseudotime", "leiden_totalVI_res0.6", "Pseudotime_bin"))

# Bind CD4 and CD4 so that colors are normalized together
n_bins_4 <- dim(bindata4)[1]
n_bins_8 <- dim(bindata8)[1]
bindata_both_prot <- rbind(bindata4, bindata8) %>% # concatenate bins x genes for CD4, CD8
  mutate(across(where(is.numeric), scale)) %>% # standard scale for each gene
  mutate(across(where(is.numeric), ~ Winsorize(.x, probs = c(0.05, 0.95)))) # Winsorize for each gene

# Separate into CD4 and CD8
norm_CD4 <- bindata_both_prot[1:n_bins_4, ]
norm_CD8 <- bindata_both_prot[(n_bins_4+1):(n_bins_4+n_bins_8), ]

# Now summarize by bin separately for each lineage
norm_CD4 <- norm_CD4 %>% 
  group_by(Bins) %>%
  summarize_all(mean)
norm_CD8 <- norm_CD8 %>% 
  group_by(Bins) %>%
  summarize_all(mean)

# Make a matrix and add row/column names
CD4_mat <- as.matrix(norm_CD4[, 2:dim(norm_CD4)[2]])
row.names(CD4_mat) <- norm_CD4$Bins
CD4_mat <- t(CD4_mat)

CD8_mat <- as.matrix(norm_CD8[, 2:dim(norm_CD8)[2]])
row.names(CD8_mat) <- norm_CD8$Bins
CD8_mat <- t(CD8_mat)

# Matrix of data to plot
both_prot_mat <- cbind(CD4_mat, CD8_mat)
colnames(both_prot_mat) <- seq(1:dim(both_prot_mat)[2])
gap_loc <- dim(CD4_mat)[2] # Where is the gap between CD4/CD8
  
# Find pseudotime values to use for annotations
bin_ranges4 <- tibble(bin_ranges = colnames(CD4_mat)) %>%
  mutate(Pseudotime = as.numeric(str_replace(str_replace(bin_ranges, "[\\(\\[]", ""), ",.*", ""))) %>% # remove either the first ( or the first [ of the bin range
  mutate(Time_bin = cut(Pseudotime, breaks = seq(0,16, by = 2), include.lowest = TRUE, labels = str_c(as.character(seq(0, 7)))))
bin_ranges4[is.na(bin_ranges4$Time_bin), "Time_bin"] <- "7" # Assign cells > 16 to the last bin 

bin_ranges8 <- tibble(bin_ranges = colnames(CD8_mat)) %>%
  mutate(Pseudotime = as.numeric(str_replace(str_replace(bin_ranges, "[\\(\\[]", ""), ",.*", ""))) %>% # remove either the first ( or the first [ of the bin range
  mutate(Time_bin = cut(Pseudotime, breaks = seq(0,16, by = 2), include.lowest = TRUE, labels = str_c(as.character(seq(0, 7)))))
bin_ranges8[is.na(bin_ranges8$Time_bin), "Time_bin"] <- "7" 

# Annotations df
annotation_df4 <- data.frame(Pseudotime = bin_ranges4[, 2])
row.names(annotation_df4) <- bin_ranges4$bin_ranges
annotation_df8 <- data.frame(Pseudotime = bin_ranges8[, 2])
row.names(annotation_df8) <- bin_ranges8$bin_ranges
annotation_df_both <- rbind(annotation_df4, annotation_df8) # Pseudotime value
annotation_df_both[["Lineage"]] <- c(rep("CD4", dim(annotation_df4)[1]), rep("CD8", dim(annotation_df8)[1])) # Lineage
row.names(annotation_df_both) <- seq(1:dim(both_prot_mat)[2])

# Annotation colors
time_bin_colors <-  brewer.pal(8, "Spectral")
names(time_bin_colors) <- as.character(seq(0, 7))
ann_colors_both <- list(
                       #Time_bin = time_bin_colors,
                       Pseudotime = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100),
                       Lineage = c("CD4" = brewer.pal(3, "Set1")[1],
                                   "CD8" = brewer.pal(3, "Set1")[2]))
# Order of features
max_time4 <- apply(CD4_mat, 1, which.max)

# Row names (genes)
gene_names_rows <- tibble(Gene = row.names(both_prot_mat[order(max_time4),])) # Get current row names for all genes
```

```{r Select DE genes to label in the heatmap}
# Label the top DE genes over time that are shared across lineages
d8_top_temporal <- de_features8 %>%
  select(Gene = X1, everything()) %>%
  filter(lfc_median > 0.2) %>%
  filter(bayes_factor > 2) %>%
  filter(non_zeros_proportion1 > 0.2) %>%
  filter(!(str_detect(Gene, "ADT_"))) %>%
  group_by(clusters) %>%
  arrange(desc(lfc_median)) %>%
  slice_head(n = 8) %>%
  ungroup() %>%
  select(Gene) %>%
  pull() %>%
  unique()

d4_top_temporal <- de_features4 %>%
  select(Gene = X1, everything()) %>%
  filter(lfc_median > 0.2) %>%
  filter(bayes_factor > 2) %>%
  filter(non_zeros_proportion1 > 0.2) %>%
  filter(!(str_detect(Gene, "ADT_"))) %>%
  group_by(clusters) %>%
  arrange(desc(lfc_median)) %>%
  slice_head(n = 8) %>%
  ungroup() %>%
  select(Gene) %>%
  pull() %>%
  unique()

toptemporaldeboth <- intersect(d4_top_temporal, d8_top_temporal)

# Generate row names, add as "labels_row" to plot in pheatmap instead of row names
gene_names_rows_select <- gene_names_rows %>%
  mutate(labels_individual = case_when(Gene %in% toptemporaldeboth ~ Gene,
                                       TRUE ~ "")) %>%
  mutate(row_number = row_number()) %>%
  mutate(group_number = cut_interval(row_number, 15)) # make 15 groups genes of equal size

# Make labels from groups
group_labels <- gene_names_rows_select %>%
  filter(labels_individual != "") %>%
  group_by(group_number) %>%
  summarize(labels_group = str_c(labels_individual, collapse = ", ")) %>%
  mutate(label_position = str_replace(labels_group, ",.*", ""))
group_labels

# Make vector of labels
group_labels_vec <- group_labels$labels_group
names(group_labels_vec) <- group_labels$label_position

# Make labels
gene_names_rows_all <- gene_names_rows_select %>%
    mutate(labels_grouped = case_when(Gene %in% group_labels$label_position ~ Gene,
                                       TRUE ~ "")) %>%
  mutate(labels_grouped_multi = str_replace_all(labels_grouped, group_labels_vec))
```

```{r Plot the multi-label heatmap}
# # Plot to save
# pheatmap(both_prot_mat[order(max_time4),],
#        cluster_cols = FALSE,
#        cluster_rows = FALSE,
#        border_color = NA,
#        show_colnames = FALSE,
#        color = colorRampPalette(viridis(10))(100),
#        show_rownames = TRUE,
#        labels_row = gene_names_rows_all %>% select(labels_grouped_multi) %>% pull(),
#        annotation_col = annotation_df_both,
#        annotation_colors = ann_colors_both,
#        cellwidth = .5,
#        cellheight = .15, 
#        treeheight_row = 0,
#        filename = str_c("figures/", "heatmap_de_genes_labels", ".png"),
#        gaps_col = c(gap_loc),
#        fontsize = 8,
#       fontsize_row = 6.5)

# Plot to view
pheatmap(both_prot_mat[order(max_time4),], 
       cluster_cols = FALSE, 
       cluster_rows = FALSE, 
       border_color = NA, 
       show_colnames = FALSE, 
       color = colorRampPalette(viridis(10))(100), 
       show_rownames = TRUE,
       labels_row = gene_names_rows_all %>% select(labels_grouped_multi) %>% pull(),
       annotation_col = annotation_df_both,
       annotation_colors = ann_colors_both,
       cellwidth = .35, 
       cellheight = .15, 
       treeheight_row = 0,
       gaps_col = c(gap_loc),
       fontsize = 8,
       fontsize_row = 7)
```

# Bin pseudotime for visualization and DE

```{r Make pseudotime bins}
# pseudotime_data contains all cells used in Slingshot
time_bins <- pseudotime_data %>%
  mutate(Pseudotime_bin = cut(mean_pseudotime, breaks = seq(0,16, by = 2), include.lowest = TRUE,
                              labels = str_c("Time", as.character(seq(0, 14, by = 2)))))

# Assign cells > 16 to the last bin
time_bins[is.na(time_bins$Pseudotime_bin), "Pseudotime_bin"] <- "Time14" # This is exactly 20 cells
# Note that these bins are used for visualization. DE testing uses smaller time bins
```

```{r Save bins for totalVI DE}
# Metadata: for all cells: Barcode, Pseudotime, Pseudotime_bin, Lineage, Genotype
totalVI_bins <- totalVI_meta %>%
  select(c(Barcode, Sample, Genotype, Experiment, Location, Mouse)) %>%
  left_join(time_bins %>% select(c("Barcode", "mean_pseudotime", "Lineage_by_genotype", "Lineage_by_genotypeSlingshot", "Pseudotime_bin")), by = "Barcode") %>%
  replace_na(list(Lineage_by_genotype = "unassigned", Lineage_by_genotypeSlingshot = "unassigned")) # replace NA values in lineages by "unassigned"
totalVI_bins

write_csv(totalVI_bins, "pseudotime_slingshot_binned_20200913.csv")
totalVI_bins <- read_csv("pseudotime_slingshot_binned_20200913.csv")
```

# Run Slingshot on full latent space

```{r Run slingshot on high dimensions}
# Filter latent space to run on just selected cells
totalVI_latent_filt <- totalvi_latent[totalVI_clusters_filt$Barcode, ]
totalVI_umap_filt <- totalvi_umap[totalVI_clusters_filt$Barcode, ]

lin <- getLineages(
  totalVI_latent_filt, 
  totalVI_clusters_filt$slingshot_clusters,
  start.clus = "1",
  end.clus = c("0", "3")
  )
lin 
```

```{r Run Slingshot get curves on high dimensions}
crv <- getCurves(lin, approx_points = 200)
```

```{r Plot results for high dimension Slingshot}
# Plot pseudotime ordering of Slingshot in high dimensions onto 2D UMAP plot
pseudotime <- as_tibble(slingPseudotime(crv)) %>%
  mutate(difference = abs(curve1 - curve2)) 
pseudotime <- pseudotime %>%
  mutate(mean_pseudotime = rowMeans(select(pseudotime, c(curve1, curve2)), na.rm = TRUE)) %>%
  mutate(UMAP1 = totalVI_umap_filt[,1]) %>%
  mutate(UMAP2 = totalVI_umap_filt[,2])

ggplot() +
  theme_classic() +
  geom_point(pseudotime, mapping = aes(x = UMAP1, y = UMAP2, color = mean_pseudotime), stroke  = 0, size = .8) +
  scale_color_gradientn(colors = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)) + 
  coord_fixed(ratio = 1) +
  labs(color = "Pseudotime") +
  labs(title = "Slingshot (full latent space)")
```

```{r Plot each lineage for high dimension Slingshot}
ggplot() +
  geom_point(pseudotime, mapping = aes(x = UMAP1, y = UMAP2), color = "grey75", stroke  = 0, size = .8) +
  theme_classic() +
  geom_point(pseudotime, mapping = aes(x = UMAP1, y = UMAP2, color = curve1), stroke  = 0, size = .8) +
  scale_color_gradientn(colors = brewer.spectral(100)) +
  coord_fixed(ratio = 1) +
  labs(color = "Pseudotime")

ggplot() +
  geom_point(pseudotime, mapping = aes(x = UMAP1, y = UMAP2), color = "grey75", stroke  = 0, size = .8) + 
  theme_classic() +
  geom_point(pseudotime, mapping = aes(x = UMAP1, y = UMAP2, color = curve2), stroke  = 0, size = .8) +
  scale_color_gradientn(colors = brewer.spectral(100)) +
  coord_fixed(ratio = 1) +
  labs(color = "Pseudotime")
```

# Compare pseudotime orderings (2D vs full latent space)

```{r Plot pseudotime orderings from 2D and 20D}
# 2D
pseudotime_data <- read_csv("pseudotime_slingshot_2020913.csv", col_types = cols(Lineage_by_genotype = col_character(), 
                                                                                  Lineage_by_genotypeSlingshot = col_character()))

# 20D (full latent space)
pseudotime_all <- pseudotime %>%
  mutate(Barcode = totalVI_clusters_filt$Barcode) %>%
  left_join(pseudotime_data, by = "Barcode", suffix = c("_20D", "_2D"))

cor(pseudotime_all$mean_pseudotime_2D, pseudotime_all$mean_pseudotime_20D, method = "spearman") #  0.9747765
cor(pseudotime_all$mean_pseudotime_2D, pseudotime_all$mean_pseudotime_20D, method = "pearson") # 0.9778185

ggplot(pseudotime_all, aes(x = mean_pseudotime_2D, y = mean_pseudotime_20D, color = mean_pseudotime_2D)) +
  geom_point(stroke = 0, size = 0.5) +
  annotate("text", label = "Pearson R: 0.978", x = 3.5, y = 0.8) +
  scale_color_gradientn(colors = colorRampPalette(brewer.pal(11,'Spectral')[-6])(100)) +
  labs(x = "Pseudotime (2D)", y = "Pseudotime (20D)", color = "Pseudotime (2D)") +
  theme_classic() +
  theme(legend.position = "bottom", aspect.ratio = 1/1)
ggsave("figures/pseudotime_2d_20d_comparison.png", bg = "transparent", height = 4.5, width = 4, dpi = 1000)
```