---
title: "Thymus DE Analysis"
output: html_notebook
---

Zoë Steier

* Analyze DE results over pseudotime.
* Transcription factor enrichment analysis.

# Load packages

```{r Load packages}
library(tidyverse)
library(RColorBrewer)

# For ChEA
library(httr)
library(jsonlite)
library(readxl)
library(openxlsx)

library(pheatmap)
library(scales)
library(pals)

packageVersion("httr")
packageVersion("jsonlite")
packageVersion("readxl")
packageVersion("openxlsx")
```

# Load DE results and metadata

```{r Load DE Results}
DE_smallbins_lin <- read_csv("DE_results/DE_withinPseudotime_CD4_vs_CD8_smallbin.csv") %>%
  select(Gene = "X1", everything())
```

```{r Load expression data}
totalvi_path <- "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/totalVI_thymus/thymus111_allbatches_stable_posselecting/totalVI_results/"
totalVI_genes <- read_csv(gzfile(str_c(totalvi_path, "totalVI_genes.csv.gz"))) %>%
  select(Gene = `0`) %>%
  pull()

# Load gene expression data
denoised_genes <- read_csv(gzfile(str_c(totalvi_path, "denoised_genes.csv.gz")))
totalVI_genes <- read_csv(gzfile(str_c(totalvi_path, "totalVI_genes.csv.gz")))
obs <- read_csv(gzfile(str_c(totalvi_path, "obs.csv.gz")))

meta_csv <- read_csv("/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/metadata/Metadata_SeqStats_totalSeq_experiments.csv")
meta_thymus <- meta_csv %>%
  filter(Tissue == "thymus") %>%
  filter(TotalSeq_panel == "ADT111") # 19 batches in total
meta_thymus <- meta_thymus %>%
  mutate(Batch = seq(1, dim(meta_thymus)[1]) - 1) # rename batches starting at 0

# Filtered data
filt_path <- "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/Annotation_thymus/"
obs_filtered <- read_csv(gzfile(str_c(filt_path, "adata_positiveselection_filt_obs.csv.gz")))
filtered_cells <- obs_filtered$X1

denoised_genes <- denoised_genes %>%
  select(-c(X1))
colnames(denoised_genes) <- totalVI_genes[, 2][[1]]
denoised_genes <- denoised_genes %>%
  mutate(Barcode = obs$X1)

pseudotime_data <- read_csv("/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/Slingshot_thymus/pseudotime_slingshot_2020913.csv", 
                            col_types = cols(Lineage_by_genotype = col_character(), Lineage_by_genotypeSlingshot = col_character()))
```

# DE over pseudotime (timeline)
## Make small time bins for DE testing

```{r Load/save results with small bins}
time_bins <- pseudotime_data %>%
  mutate(Pseudotime_bin = cut(mean_pseudotime, breaks = seq(0,15, by = 1), include.lowest = TRUE,
                              labels = str_c(as.character(seq(0, 14)))))
# Assign cells > 16 to the last bin
time_bins[is.na(time_bins$Pseudotime_bin), "Pseudotime_bin"] <- "14" # combine last two bins, since only ~100 CD8 cells fall in final bin

as.data.frame.matrix(table(time_bins$Pseudotime_bin, time_bins$Genotype))
table(time_bins$Pseudotime_bin, time_bins$Lineage_by_genotypeSlingshot)

# Read in pseudotime data from all cells
pseudotime_binned <- read_csv("/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/Slingshot_thymus/pseudotime_slingshot_binned_20200913.csv")

time_bins_allcells <- pseudotime_binned %>% # keep cells in order from all cells
  select(Barcode, Lineage_by_genotypeSlingshot) %>%
  left_join(time_bins %>% select(-c("Lineage_by_genotypeSlingshot")), by = "Barcode") %>%
  select(Barcode, mean_pseudotime, Pseudotime_bin, Lineage_by_genotypeSlingshot, everything())

write_csv(time_bins_allcells, "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/Slingshot_thymus/pseudotime_slingshot_binned_20210202.csv")
```

## DE over pseudotime with bin ranges

```{r Timeline data to plot}
DE_timeline <- DE_smallbins_lin %>%
  filter(non_zeros_proportion1 > 0.05 | non_zeros_proportion2 > 0.05) %>% # must be expressed at least in one group
  filter(!(str_detect(Gene, "ADT_"))) %>%
  filter(Gene %in% c("Gata3", "Runx3", "Zbtb7b", "Cd69", "Egr1", "Cd4", "Cd8a", "CD5", "Cd5")) %>%
  mutate(Significant = case_when(
      bayes_factor > 2 &
      (((non_zeros_proportion1 > 0.05) & (lfc_median > .2)) | ((non_zeros_proportion2 > 0.05) & (lfc_median < -.2))) ~ "Significant",
      TRUE ~ "Not significant" 
    )) %>%
  select(Gene, clusters, lfc_median, bayes_factor, Significant, non_zeros_proportion1, non_zeros_proportion2, lfc_std)

DE_timeline_proteins <- DE_smallbins_lin %>%
  filter((str_detect(Gene, "ADT_"))) %>%
  filter(Gene %in% c("ADT_CD4_A0001", "ADT_CD8a_A0002", "ADT_CD5_A0111")) %>%
  mutate(Significant = case_when(
      bayes_factor > 1 &
      (((lfc_median > .2)) | ((lfc_median < -.2))) ~ "Significant",
      TRUE ~ "Not significant" 
    )) %>%
  mutate(Gene = recode(Gene, ADT_CD4_A0001 = "CD4", ADT_CD8a_A0002 = "CD8a", ADT_CD5_A0111 = "CD5")) %>%
  select(Gene, clusters, lfc_median, bayes_factor, Significant, lfc_std) 

DE_timeline_all <- rbind(DE_timeline %>% select(-c(non_zeros_proportion1, non_zeros_proportion2)), DE_timeline_proteins) %>%
  select(Feature = Gene, everything())
```

```{r Add bin ranges to data}
DE_timeline_all <- DE_timeline_all %>%
  mutate(Bin_ranges = factor(recode(clusters, `0` = "0-1",
                             `1` = "1-2",
                             `2` = "2-3",
                             `3` = "3-4",
                             `4` = "4-5",
                             `5` = "5-6",
                             `6` = "6-7",
                             `7` = "7-8",
                             `8` = "8-9",
                             `9` = "9-10",
                             `10` = "10-11",
                             `11` = "11-12",
                             `12` = "12-13",
                             `13` = "13-14",
                             `14` = "14-16"), levels = c("0-1", "1-2","2-3","3-4","4-5","5-6","6-7","7-8","8-9","9-10","10-11","11-12","12-13","13-14","14-16")))

bin_ranges_labels <- c(`0` = "0-1",
                             `1` = "1-2",
                             `2` = "2-3",
                             `3` = "3-4",
                             `4` = "4-5",
                             `5` = "5-6",
                             `6` = "6-7",
                             `7` = "7-8",
                             `8` = "8-9",
                             `9` = "9-10",
                             `10` = "10-11",
                             `11` = "11-12",
                             `12` = "12-13",
                             `13` = "13-14",
                             `14` = "14-16")
```

```{r Plot DE over pseudotime timeline with bin ranges}
# Plot coreceptors
ggplot() +
  geom_errorbar(data = DE_timeline_all %>% filter(Feature %in% c("Cd4", "CD4", "Cd8a", "CD8a")),
          aes(x = clusters, y = lfc_median, ymin = lfc_median - lfc_std, ymax = lfc_median + lfc_std, width = .2), color = "gray", alpha = .4, size = .2) +
  geom_errorbar(data = DE_timeline_all %>% filter(Feature %in% c("Cd4", "CD4", "Cd8a", "CD8a"), Significant == "Significant"),
          aes(x = clusters, y = lfc_median, ymin = lfc_median - lfc_std, ymax = lfc_median + lfc_std, color = Feature, linetype = Feature, width = .2), size = .2) +
  geom_line(data = DE_timeline_all %>% filter(Feature %in% c("Cd4", "CD4", "Cd8a", "CD8a")), 
            aes(x = clusters, y = lfc_median, color = Feature, linetype = Feature), size = .2) +
  geom_point(data = DE_timeline_all %>% filter(Feature %in% c("Cd4", "CD4", "Cd8a", "CD8a")),
             aes(x = clusters, y = lfc_median, size = bayes_factor), color = "gray", alpha = 0.4) +
  scale_size_area(limits = c(0.2,5), max_size = 4) +
  geom_point(data = DE_timeline_all %>% filter(Significant == "Significant", Feature %in% c("Cd4", "CD4", "Cd8a", "CD8a")),
             aes(x = clusters, y = lfc_median, color = Feature, size = bayes_factor, shape = Feature)) +
  scale_shape_manual(values = c("Cd4" = 16, "CD4" = 21, "Cd8a" = 16, "CD8a" = 1, "Gata3" = 16, "Runx3" = 16, "Zbtb7b" = 16, "Cd69" = 16, "Egr1" = 16)) +
  scale_color_manual(values = c("Cd4" = "#F8766D", "CD4" = "#F8766D", "Cd8a" = "#00B6EB", "CD8a" = "#00B6EB", "Gata3" = "#A58AFF", "Runx3" = "#00C094", "Zbtb7b" = "#FB61D7", "Cd69" = "#C49A00", "Egr1" = "#53B400"),
                     breaks = c("Cd4", "CD4", "Cd8a", "CD8a", "Gata3", "Runx3", "Zbtb7b", "Cd69", "Egr1")) +
  scale_linetype_manual(values = c("Cd4" = "longdash", "CD4" = "solid", "Cd8a" = "longdash", "CD8a" = "solid", "Gata3" = "longdash", "Runx3" = "longdash", "Zbtb7b" = "longdash", "Cd69" = "longdash", "Egr1" = "longdash"),
                     breaks = c("Cd4", "CD4", "Cd8a", "CD8a", "Gata3", "Runx3", "Zbtb7b", "Cd69", "Egr1")) +
  theme_classic() +
  scale_x_continuous(breaks = seq(0, 14, 1), minor_breaks = NULL, labels = bin_ranges_labels) +
  theme(axis.text.x = element_text(angle = 45, vjust = .5)) +
  geom_hline(yintercept = 0) +
  labs(x = "Pseudotime", y = "Log fold change", size = "Bayes factor", color = "Feature", linetype = "Feature", shape = "Feature") 
ggsave("figures/DEtiming_coreceptors.pdf", bg = "transparent", height = 3.5, width = 5)

# Plot master regulator TFs
ggplot() +
  geom_errorbar(data = DE_timeline_all %>% filter(Feature %in% c("Gata3", "Runx3", "Zbtb7b")),
          aes(x = clusters, y = lfc_median, ymin = lfc_median - lfc_std, ymax = lfc_median + lfc_std, width = .2), color = "gray", alpha = .4, size = .2) +
  geom_errorbar(data = DE_timeline_all %>% filter(Feature %in% c("Gata3", "Runx3", "Zbtb7b"), Significant == "Significant"),
          aes(x = clusters, y = lfc_median, ymin = lfc_median - lfc_std, ymax = lfc_median + lfc_std, color = Feature, linetype = Feature, width = .2), size = .2) +
  geom_line(data = DE_timeline_all %>% filter(Feature %in% c("Gata3", "Runx3", "Zbtb7b")), 
          aes(x = clusters, y = lfc_median, color = Feature, linetype = Feature, group = Feature), size = .2, alpha = 1) +
  geom_point(data = DE_timeline_all %>% filter(Feature %in% c("Gata3", "Runx3", "Zbtb7b")),
             aes(x = clusters, y = lfc_median, size = bayes_factor), color = "gray", alpha = 0.4) +
  geom_point(data = DE_timeline_all %>% filter(Significant == "Significant", Feature %in% c("Gata3", "Runx3", "Zbtb7b")),
             aes(x = clusters, y = lfc_median, color = Feature, size = bayes_factor, shape = Feature), alpha = 1) +
  geom_point(data = DE_timeline_all %>% filter(Significant == "Significant", Feature %in% c("Zbtb7b")), # Make Zbtb7b visible
             aes(x = clusters, y = lfc_median, color = Feature, size = bayes_factor, shape = Feature), alpha = 1) +
  scale_size_area(limits = c(0.2,5), max_size = 4) +
  scale_shape_manual(values = c("Cd4" = 16, "CD4" = 1, "Cd8a" = 16, "CD8a" = 1, "Gata3" = 16, "Runx3" = 16, "Zbtb7b" = 16, "Cd69" = 16, "Egr1" = 16)) +
  scale_color_manual(values = c("Cd4" = "#F8766D", "CD4" = "#F8766D", "Cd8a" = "#00B6EB", "CD8a" = "#00B6EB", "Gata3" = "#A58AFF", "Runx3" = "#00C094", "Zbtb7b" = "#FB61D7", "Cd69" = "#C49A00", "Egr1" = "#53B400"),
                     breaks = c("Cd4", "CD4", "Cd8a", "CD8a", "Gata3", "Runx3", "Zbtb7b", "Cd69", "Egr1")) +
  scale_linetype_manual(values = c("Cd4" = "longdash", "CD4" = "solid", "Cd8a" = "longdash", "CD8a" = "solid", "Gata3" = "longdash", "Runx3" = "longdash", "Zbtb7b" = "longdash", "Cd69" = "longdash", "Egr1" = "longdash"),
                        breaks = c("Cd4", "CD4", "Cd8a", "CD8a", "Gata3", "Runx3", "Zbtb7b", "Cd69", "Egr1")) +
  theme_classic() +
  scale_x_continuous(breaks = seq(0, 14, 1), minor_breaks = NULL, labels = bin_ranges_labels) +
  theme(axis.text.x = element_text(angle = 45, vjust = .5)) +
  geom_hline(yintercept = 0) +
  labs(x = "Pseudotime", y = "Log fold change", size = "Bayes factor", color = "Feature", linetype = "Feature", shape = "Feature")
ggsave("figures/DEtiming_masterregs.pdf", bg = "transparent", height = 3.5, width = 5)

# Plot TCR response genes
ggplot() +
  geom_errorbar(data = DE_timeline_all %>% filter(Feature %in% c("Cd69", "Egr1")),
        aes(x = clusters, y = lfc_median, ymin = lfc_median - lfc_std, ymax = lfc_median + lfc_std, width = .2), color = "gray", alpha = .4, size = .2) +
  geom_errorbar(data = DE_timeline_all %>% filter(Feature %in% c("Cd69", "Egr1"), Significant == "Significant"),
        aes(x = clusters, y = lfc_median, ymin = lfc_median - lfc_std, ymax = lfc_median + lfc_std, color = Feature, linetype = Feature, width = .2), size = .2) +
  geom_line(data = DE_timeline_all %>% filter(Feature %in% c("Cd69", "Egr1")), 
        aes(x = clusters, y = lfc_median, color = Feature, linetype = Feature), size = .2, alpha = 1) +
  geom_point(data = DE_timeline_all %>% filter(Feature %in% c("Cd69", "Egr1")),
        aes(x = clusters, y = lfc_median, size = bayes_factor), color = "gray", alpha = 0.4) +
  geom_point(data = DE_timeline_all %>% filter(Significant == "Significant", Feature %in% c("Cd69", "Egr1")),
             aes(x = clusters, y = lfc_median, color = Feature, shape = Feature, size = bayes_factor), alpha = 1) +
  scale_size_area(limits = c(0.2,5), max_size = 4) +
  scale_shape_manual(values = c("Cd4" = 16, "CD4" = 1, "Cd8a" = 16, "CD8a" = 1, "Gata3" = 16, "Runx3" = 16, "Zbtb7b" = 16, "Cd69" = 16, "Egr1" = 16),
                     breaks = c("Cd4", "CD4", "Cd8a", "CD8a", "Gata3", "Runx3", "Zbtb7b", "Cd69", "Egr1")) +
  scale_color_manual(values = c("Cd4" = "#F8766D", "CD4" = "#F8766D", "Cd8a" = "#00B6EB", "CD8a" = "#00B6EB", "Gata3" = "#A58AFF", "Runx3" = "#00C094", "Zbtb7b" = "#FB61D7", "Cd69" = "#C49A00", "Egr1" = "#53B400"),
                     breaks = c("Cd4", "CD4", "Cd8a", "CD8a", "Gata3", "Runx3", "Zbtb7b", "Cd69", "Egr1")) +
  scale_linetype_manual(values = c("Cd4" = "longdash", "CD4" = "solid", "Cd8a" = "longdash", "CD8a" = "solid", "Gata3" = "longdash", "Runx3" = "longdash", "Zbtb7b" = "longdash", "Cd69" = "longdash", "Egr1" = "longdash"),
                        breaks = c("Cd4", "CD4", "Cd8a", "CD8a", "Gata3", "Runx3", "Zbtb7b", "Cd69", "Egr1")) +
  theme_classic() +
  scale_x_continuous(breaks = seq(0, 14, 1), minor_breaks = NULL, labels = bin_ranges_labels) +
  theme(axis.text.x = element_text(angle = 45, vjust = .5)) +
  geom_hline(yintercept = 0) +
  labs(x = "Pseudotime", y = "Log fold change", color = "Feature", linetype = "Feature", shape = "Feature", size = "Bayes factor") +
  guides(color = guide_legend(order = 1), linetype = guide_legend(order = 1), shape = guide_legend(order = 1), size = guide_legend(order = 2)) # changes order of multiple guides to match
ggsave("figures/DEtiming_tcrresponse.pdf", bg = "transparent", height = 3.5, width = 5)
```

## Number of DE genes over time for small bins

```{r Between lineage DE small bins}
# See how the number of DE genes and proteins increases over pseudotime
DE_betweenlineage <- DE_smallbins_lin

# RNA (DE in either direction)
DE_RNA_betweenLineage <- c()
DErna4 <- c()
DErna8 <- c()
for (cluster in unique(DE_betweenlineage$clusters)) {
  DE_bylin <- DE_betweenlineage %>%
    filter(bayes_factor > 2) %>% 
    filter( ((non_zeros_proportion1 > 0.05) & (lfc_median > .2)) | ((non_zeros_proportion2 > 0.05) & (lfc_median < -.2))) %>%
    filter(!(str_detect(Gene, "ADT_"))) %>%
    filter(clusters == cluster)
  number_DE <- dim(DE_bylin)[1]
  number_4 <- dim(DE_bylin %>% filter(lfc_median > 0.2))[1]
  number_8 <- dim(DE_bylin %>% filter(lfc_median < -0.2))[1]
  
  DE_RNA_betweenLineage <- c(DE_RNA_betweenLineage, number_DE)
  DErna4 <- c(DErna4, number_4)
  DErna8 <- c(DErna8, number_8)
}

# Proteins
DE_protein_betweenLineage <- c()
DEprot4 <- c()
DEprot8 <- c()
for (cluster in unique(DE_betweenlineage$clusters)) {
  DE_bylin <- DE_betweenlineage %>%
    filter(bayes_factor > 1) %>% 
    filter(clusters == cluster) %>%
    filter((str_detect(Gene, "ADT_"))) %>%
    filter(abs(lfc_median) > 0.2)
  number_DE <- dim(DE_bylin)[1]
  number_4 <- dim(DE_bylin %>% filter(lfc_median > 0.2))[1]
  number_8 <- dim(DE_bylin %>% filter(lfc_median < -0.2))[1]
  
  DE_protein_betweenLineage <- c(DE_protein_betweenLineage, number_DE)
  DEprot4 <- c(DEprot4, number_4)
  DEprot8 <- c(DEprot8, number_8)
}

# Collect results
DEnum_byTime_rna <- tibble(time = unique(DE_betweenlineage$clusters), RNA = DE_RNA_betweenLineage, CD4 = DErna4, CD8 = DErna8) %>%
  pivot_longer(cols = "RNA", names_to = "Feature", values_to = "Number_DE") 
DEnum_byTime_prot <- tibble(time = unique(DE_betweenlineage$clusters), Protein = DE_protein_betweenLineage, CD4 = DEprot4, CD8 = DEprot8) %>%
    pivot_longer(cols = "Protein", names_to = "Feature", values_to = "Number_DE")
DE_byTime <- rbind(DEnum_byTime_rna, DEnum_byTime_prot) 
DE_byTime$Feature <- factor(DE_byTime$Feature, levels = c("RNA", "Protein"))
DE_byTime$time <- factor(DE_byTime$time)
  
new_theme <- theme(
              panel.background = element_blank(),panel.grid.major = element_blank(), panel.grid.minor = element_blank(),axis.line = element_line(colour = "black"),
              strip.background = element_blank(),
              axis.text.x=element_text(angle = 45, vjust = .5)
        )

# Total number DE
ggplot(DE_byTime, aes(fill = time, y = Number_DE, x = time)) +
  geom_bar(stat = "identity") +
  scale_fill_manual(values = colorRampPalette(brewer.pal(11,'Spectral')[-6])(15), labels = bin_ranges_labels) +
  facet_wrap(~Feature, nrow = 1, scales = "free_y") +
  scale_x_discrete(labels = bin_ranges_labels) +
  new_theme +
  labs(x = "Pseudotime", y = "Differentially expressed features", fill = "Pseudotime")
ggsave("figures/DE_number_over_time.pdf", bg = "transparent", height = 4.5, width = 6) # height 4
```

# ChEA3 small bins DE between lineages
## CD4 TF enrichment

```{r Set up DE data for TF enrichment}
DE_betweenlineage <- DE_smallbins_lin

expressed_genes <- unique(c((DE_betweenlineage %>% filter(!(str_detect(Gene, "ADT_"))))$Gene))
length(expressed_genes) # 5108 (totalVI gene set)
```

```{r Run ChEA3 for CD4s}
# Find DE gene list
gene_lists = list()
for (cluster in unique(DE_betweenlineage$clusters)) {
  name <- str_c("CD4_bin", as.character(cluster))
  de_list <- (DE_betweenlineage %>% 
                              filter(bayes_factor > 2.0) %>% 
                              filter(clusters == cluster) %>% 
                              filter(non_zeros_proportion1 > 0.05) %>% 
                              filter(lfc_median > 0.2) %>% 
                              filter(!str_detect(Gene, "ADT_")))$Gene
  gene_lists[[name]] <- de_list
}

# Query ChEA database
url = "https://amp.pharm.mssm.edu/chea3/api/enrich/"
encode = "json"

ChEA_results <- list()
for (list_num in seq(1, length(gene_lists))) {
  query <- names(gene_lists)[list_num]
  gene_set <- (gene_lists[[list_num]]) # must be a character, not list
  
  if (length(gene_set) > 1) { # gene list must contain more than 1 gene
  
    payload = list(query_name = query, gene_set = gene_set)
    
    #POST to ChEA3 server
    response = POST(url = url, body = payload, encode = encode)
    json = content(response, "text")
    
    # Results as list of R dataframes
    results = fromJSON(json)
    results <- as_tibble(results[["Integrated--meanRank"]]) %>%
      mutate(TF = (str_to_title(TF))) %>%
      filter(TF %in% expressed_genes)         # filter results by whether the gene is expressed in this data
    
    ChEA_results[[query]] <- results
  }
}

cd4_de_list_lengths <- lengths(gene_lists)
```

```{r Save CD4 ChEA3 results}
write.xlsx(ChEA_results, file = "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/DE_thymus/ChEA3_results/CD4_TFenrichment_smallbins.xlsx")

CD4_genes <- tibble(Time_bin = names(gene_lists))
CD4_genes$DE_genes <- ""
for (i in seq(length(gene_lists))) {
  CD4_genes[i, "DE_genes"] <- str_c(unlist(gene_lists[i]), collapse = ", ")
}
write_csv(CD4_genes, "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/DE_thymus/ChEA3_results/CD4_DE_genes_smallbins.csv")
```

## CD8 TF enrichment

```{r Run ChEA3 for CD8s}
# Find DE gene list
gene_lists = list()
for (cluster in unique(DE_betweenlineage$clusters)) {
  name <- str_c("CD8_bin", as.character(cluster))
  de_list <- (DE_betweenlineage %>% 
                              filter(bayes_factor > 2.0) %>% 
                              filter(clusters == cluster) %>% 
                              filter(non_zeros_proportion2 > 0.05) %>% 
                              filter(lfc_median < -0.2) %>% 
                              filter(!str_detect(Gene, "ADT_")))$Gene
  gene_lists[[name]] <- de_list
}

# Query ChEA database
url = "https://amp.pharm.mssm.edu/chea3/api/enrich/"
encode = "json"

ChEA_results <- list()
for (list_num in seq(1, length(gene_lists))) {
  query <- names(gene_lists)[list_num]
  gene_set <- (gene_lists[[list_num]]) # must be a character, not list
  if (length(gene_set) > 1) { 
    payload = list(query_name = query, gene_set = gene_set)
    
    #POST to ChEA3 server
    response = POST(url = url, body = payload, encode = encode)
    json = content(response, "text")
    
    # Results as list of R dataframes
    results = fromJSON(json)
    results <- as_tibble(results[["Integrated--meanRank"]]) %>%
      mutate(TF = (str_to_title(TF))) %>%
      filter(TF %in% expressed_genes)         # filter results by whether the gene is expressed in this data
    
    ChEA_results[[query]] <- results
  }
}

cd8_de_list_lengths <- lengths(gene_lists)
```
```{r Save CD8 ChEA3 results}
write.xlsx(ChEA_results, file = "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/DE_thymus/ChEA3_results/CD8_TFenrichment_smallbins.xlsx")

CD8_genes <- tibble(Time_bin = names(gene_lists))
CD8_genes$DE_genes <- ""
for (i in seq(length(gene_lists))) {
  CD8_genes[i, "DE_genes"] <- str_c(unlist(gene_lists[i]), collapse = ", ")
}
write_csv(CD8_genes, "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/DE_thymus/ChEA3_results/CD8_DE_genes_smallbins.csv")
```

# ChEA3 on small bin clustered genes
## Plot clustered genes

```{r Load gene clusters}
cd4_gene_clusters <- read_csv("DE_results/clusteredDE_CD4_genes_smallbins.csv")
cd8_gene_clusters <- read_csv("DE_results/clusteredDE_CD8_genes_smallbins.csv")
temporal_gene_clusters <- read_csv("DE_results/clusteredDE_temporal_genes_smallbins.csv")

cd4_gene_clusters$Cluster <- as.factor(cd4_gene_clusters$Cluster)
cd8_gene_clusters$Cluster <- as.factor(cd8_gene_clusters$Cluster)
temporal_gene_clusters$Cluster <- as.factor(temporal_gene_clusters$Cluster)
```

```{r Merge genes and meta}
# Scaling to 98th percentile so the scale isn't biased by high outliers
scale_98 <- function(x, na.rm = TRUE) {rescale(x, to = c(0,1), from = c(min(x), quantile(x, .98)))}

# Add metadata and scale
denoised_genes_annotated <- denoised_genes %>%
  left_join(pseudotime_data %>% select(Barcode, mean_pseudotime, Lineage_by_genotypeSlingshot, Genotype), by = "Barcode") %>%
  filter(Lineage_by_genotypeSlingshot %in% c("CD4", "CD8")) %>%
  select(all_of(totalVI_genes[, 2][[1]]), Lineage_by_genotypeSlingshot, mean_pseudotime, Genotype) %>%
  mutate_at(totalVI_genes[, 2][[1]], scale_98) %>%
  pivot_longer(cols = all_of(totalVI_genes[, 2][[1]]), names_to = "Gene")
```

```{r Plot CD4 clustered DE genes}
redcol <- brewer.pal(3, "Set1")[1]
bluecol <- brewer.pal(3, "Set1")[2]

# Select DE genes
CD4_genes_annotated <- denoised_genes_annotated %>%
  filter(Gene %in% cd4_gene_clusters$Gene) %>%
  left_join(cd4_gene_clusters, by = "Gene") %>%
  group_by(Cluster, mean_pseudotime, Lineage_by_genotypeSlingshot) %>% # group by cluster, time, lineage and take mean
  summarize(cluster_mean = mean(value))

# Plot by loess, facet by cluster
ggplot(CD4_genes_annotated, aes(x = mean_pseudotime, y = cluster_mean, color = Lineage_by_genotypeSlingshot)) +
  geom_smooth(method = "loess", span = .5, se = FALSE) +
  scale_color_manual(values = c(redcol, bluecol)) +
  facet_wrap("Cluster", ncol = 5) +
  labs(x = "Pseudotime", y = "Mean scaled\n denoised expression", color = "Lineage", title = "MHCII-specific gene clusters") +
  theme_classic()
ggsave("figures/DEgeneclusters_CD4.pdf", bg = "transparent", height = 2.5, width = 7.5)
```

```{r Plot CD8 clustered DE genes}
# Select DE genes
CD8_genes_annotated <- denoised_genes_annotated %>%
  filter(Gene %in% cd8_gene_clusters$Gene) %>%
  left_join(cd8_gene_clusters, by = "Gene") %>%
  group_by(Cluster, mean_pseudotime, Lineage_by_genotypeSlingshot) %>% # group by cluster, time, lineage and take mean
  summarize(cluster_mean = mean(value))

# Plot by loess, facet by cluster
ggplot(CD8_genes_annotated, aes(x = mean_pseudotime, y = cluster_mean, color = Lineage_by_genotypeSlingshot)) +
  geom_smooth(method = "loess", span = .5, se = FALSE) +
  scale_color_manual(values = c(redcol, bluecol)) +
  facet_wrap("Cluster", ncol = 5) +
  labs(x = "Pseudotime", y = "Mean scaled\n denoised expression", color = "Lineage", title = "MHCI-specific gene clusters") +
  theme_classic()
ggsave("figures/DEgeneclusters_CD8.pdf", bg = "transparent", height = 2.5, width = 7.5)
```

## TF enrichment on gene clusters

```{r TF enrichment on CD4 clusters}
# Find DE gene list
gene_lists = list()
for (cluster in levels(cd4_gene_clusters$Cluster)) {
  name <- str_c("CD4_cluster", as.character(cluster))
  de_list <- cd4_gene_clusters %>%
    filter(Cluster == cluster) %>%
    select(Gene) %>%
    pull()
  gene_lists[[name]] <- de_list
}

# Query ChEA database
url = "https://amp.pharm.mssm.edu/chea3/api/enrich/"
encode = "json"

ChEA_results <- list()
for (list_num in seq(1, length(gene_lists))) {
  query <- names(gene_lists)[list_num]
  gene_set <- (gene_lists[[list_num]]) # must be a character, not list
  
  if (length(gene_set) > 1) { # gene list must contain more than 1 gene
  
    payload = list(query_name = query, gene_set = gene_set)
    
    #POST to ChEA3 server
    response = POST(url = url, body = payload, encode = encode)
    json = content(response, "text")
    
    # Results as list of R dataframes
    results = fromJSON(json)
    results <- as_tibble(results[["Integrated--meanRank"]]) %>%
      mutate(TF = (str_to_title(TF))) %>%
      filter(TF %in% expressed_genes)    # filter results by whether the gene is expressed in this data

    ChEA_results[[query]] <- results
  }
}
```

```{r Save CD4 results by cluster}
write.xlsx(ChEA_results, file = "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/DE_thymus/ChEA3_results/CD4_TFenrichment_clusters.xlsx")

ChEA_results

# Cluster 4: Nfatc2 is top ranked, Tcf7 and Lef1 also in top 10
# Cluster 7 is Egr
```

```{r TF enrichment on CD8 clusters}
# Find DE gene list
gene_lists = list()
for (cluster in levels(cd8_gene_clusters$Cluster)) {
  name <- str_c("CD8_cluster", as.character(cluster))
  de_list <- cd8_gene_clusters %>%
    filter(Cluster == cluster) %>%
    select(Gene) %>%
    pull()
  gene_lists[[name]] <- de_list
}

# Query ChEA database
url = "https://amp.pharm.mssm.edu/chea3/api/enrich/"
encode = "json"

ChEA_results <- list()
for (list_num in seq(1, length(gene_lists))) {
  query <- names(gene_lists)[list_num]
  gene_set <- (gene_lists[[list_num]]) # must be a character, not list
  
  if (length(gene_set) > 1) { # gene list must contain more than 1 gene
  
    payload = list(query_name = query, gene_set = gene_set)
    
    #POST to ChEA3 server
    response = POST(url = url, body = payload, encode = encode)
    json = content(response, "text")
    
    #results as list of R dataframes
    results = fromJSON(json)
    results <- as_tibble(results[["Integrated--meanRank"]]) %>%
      mutate(TF = (str_to_title(TF))) %>%
      filter(TF %in% expressed_genes)   # filter results by whether the gene is expressed in this data
    
    ChEA_results[[query]] <- results
  }
}
```

```{r Save CD8 results by cluster}
write.xlsx(ChEA_results, file = "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/DE_thymus/ChEA3_results/CD8_TFenrichment_clusters.xlsx")

ChEA_results
```

# Megatable
## Create TCR signaling gene sets

```{r Find TCR pathway genes from Netpath}
TCR_pathway <- read_csv("/data/yosef2/users/zsteier/references/Signatures/NetPath_TCR.csv", col_names = FALSE) %>%
  pivot_longer(cols = everything(), values_to = "Gene") %>%
  select(Gene) %>%
  mutate(Gene = str_to_title(Gene)) %>%
  drop_na() %>%
  pull()
```

```{r Find genes upregulated by TCR pathway from Netpath}
TCRreg <- read_csv("/data/yosef2/users/zsteier/TotalSeq/20190718_NovaSeq_ZRS07/ZRS07_ZRS06_analysis/GSEA/NetPath_TCR_regulated.csv")
TCRreg <- TCRreg[17:648, c(2, 3, 7, 8)]
colnames(TCRreg) <- TCRreg[1,]
TCRreg <- TCRreg %>%
  mutate("Gene" = str_to_title(`Gene Symbol`))
TCRregup <- TCRreg[2:450,]$Gene
```

```{r Find DP activated genes}
# See Mingeneau et al., 2012
DPactivated <- read_csv("/data/yosef2/users/zsteier/references/Signatures/Mingeneau_DPactivated.csv") %>%
  pull()

# TCR signature including all sets
TCR_sig <- unique(c(TCR_pathway, TCRregup, DPactivated))
```

```{r Create TCR signaling gene set}
TCR_set <- unique(c(TCR_pathway, TCRregup))
TCR_manual <- c("Egr2", "Runx1", "Ets1", "Nfkb2", "Foxo1", "Irf4", "Elk3", "Junb", "Jund", "Tcf7") # manual list of TCR genes
TCR_set <- c(TCR_set, TCR_manual)
```

```{r Find TFs that target master regulators}
# Read Chea results
cd4_chea_filepath <- "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/DE_thymus/ChEA3_results/CD4_TFenrichment_smallbins.xlsx"
cd8_chea_filepath <- "/data/yosef2/users/zsteier/TotalSeq/20190814_BioLegend_ZRS08/analysis/DE_thymus/ChEA3_results/CD8_TFenrichment_smallbins.xlsx"

tab_names <- excel_sheets(cd4_chea_filepath)
Chea_CD4 <- reduce(lapply(tab_names, function(x) read_excel(path = cd4_chea_filepath, sheet = x)), rbind)
Targets_masterreg4 <- Chea_CD4 %>%
  mutate(Targets_masterreg = str_detect(Overlapping_Genes, "RUNX3|ZBTB7B|GATA3")) %>%
  filter(Targets_masterreg == TRUE) %>%
  select(TF) %>%
  pull() %>%
  unique()
tab_names <- excel_sheets(cd8_chea_filepath)
Chea_CD8 <- reduce(lapply(tab_names, function(x) read_excel(path = cd8_chea_filepath, sheet = x)), rbind)
Targets_masterreg8 <- Chea_CD8 %>%
  mutate(Targets_masterreg = str_detect(Overlapping_Genes, "RUNX3")) %>%
  filter(Targets_masterreg == TRUE) %>%
  select(TF) %>%
  pull() %>%
  unique()

# Search overlapping gene list for master regulators, then filter for any TFs that target these
Targets_masterreg <- unique(c(Targets_masterreg4, Targets_masterreg8))
```

## Collect DE results

```{r CD4 select DE bins}
DE_CD4_table_viewing <- DE_smallbins_lin %>%
  filter(clusters %in% c(4, 5, 6, 7)) %>%
  filter(bayes_factor > 2) %>% 
  filter(non_zeros_proportion1 > 0.05, lfc_median > .2) %>% 
  filter(!(str_detect(Gene, "ADT_"))) %>%
  select(Gene, bayes_factor, non_zeros_proportion1, non_zeros_proportion2, lfc_median, clusters) %>%
  mutate(LFC = lfc_median)

DE_CD4_table <- DE_smallbins_lin %>%
  filter(clusters %in% c(4, 5, 6, 7)) %>%
  filter(bayes_factor > 2) %>%
  filter(non_zeros_proportion1 > 0.05, lfc_median > .2) %>%
  filter(!(str_detect(Gene, "ADT_"))) %>%
  mutate(LFC = lfc_median) %>%
  select(Gene, LFC, clusters)
```

```{r CD8 select DE bins}
# CD8 DE for megatable: bins 5-8
DE_CD8_table_view <- DE_smallbins_lin %>%
  filter(clusters %in% c(5, 6, 7, 8)) %>% # 5 is first DE > 15 genes
  filter(bayes_factor > 2) %>%
  filter(non_zeros_proportion2 > 0.05, lfc_median < -.2) %>%
  filter(!(str_detect(Gene, "ADT_"))) %>%
  select(Gene, bayes_factor, non_zeros_proportion1, non_zeros_proportion2, lfc_median, clusters) %>%
  mutate(LFC = -1*lfc_median)

DE_CD8_table <- DE_smallbins_lin %>%
  filter(clusters %in% c(5, 6, 7, 8)) %>%
  filter(bayes_factor > 2) %>% 
  filter(non_zeros_proportion2 > 0.05, lfc_median < -.2) %>% 
  filter(!(str_detect(Gene, "ADT_"))) %>%
  mutate(LFC = -1*lfc_median) %>%
  select(Gene, LFC, clusters)
```

## Collect ChEA results

```{r Collect results for CD4}
TF_CD4_table <- Chea_CD4 %>%
  filter(`Query Name` %in% c("CD4_bin4", "CD4_bin5", "CD4_bin6", "CD4_bin7")) %>%
  select(Gene = TF, everything()) %>%
  mutate(Score = as.numeric(Score)) %>% # necessary for ranking to work
  mutate(Pseudotime_bin = as.integer(str_replace(`Query Name`, "CD4_bin", ""))) %>%
  left_join(DE_smallbins_lin %>% select(Gene, non_zeros_proportion1, clusters) %>% filter(clusters %in% c(4, 5, 6, 7)), by = c("Gene", "Pseudotime_bin" = "clusters")) %>%
  filter(non_zeros_proportion1 > 0.05) %>%
  select(-c("non_zeros_proportion1")) %>%
  select(-c(`Query Name`, Library, Rank, Overlapping_Genes)) %>%
  group_by(Pseudotime_bin) %>%
  mutate(Rank_bin = rank(Score)) %>%
  pivot_wider(id_cols = c("Gene"), names_from = Pseudotime_bin, values_from = c(Score, Rank_bin)) %>%
  rowwise() %>% # take mean across row of ranks
  mutate(Rank_mean = mean(c(Rank_bin_4, Rank_bin_5, Rank_bin_6), na.rm = TRUE)) %>%
  rowwise() %>%
  mutate(Score_mean = mean(c(Score_4, Score_5, Score_6), na.rm = TRUE)) %>%
  arrange(Rank_mean, Score_mean) # Rank_mean 

TF_CD4_table
```

```{r Collect results for CD8}
TF_CD8_table <- Chea_CD8 %>%
  filter(`Query Name` %in% c("CD8_bin5", "CD8_bin6", "CD8_bin7", "CD8_bin8")) %>%
  select(Gene = TF, everything()) %>%
  mutate(Score = as.numeric(Score)) %>% # necessary for ranking to work
  mutate(Pseudotime_bin = as.integer(str_replace(`Query Name`, "CD8_bin", ""))) %>%
  left_join(DE_smallbins_lin %>% select(Gene, non_zeros_proportion2, clusters) %>% filter(clusters %in% c(5, 6, 7, 8)), by = c("Gene", "Pseudotime_bin" = "clusters")) %>%
  filter(non_zeros_proportion2 > 0.05) %>%
  select(-c("non_zeros_proportion2")) %>%
  select(-c(`Query Name`, Library, Rank, Overlapping_Genes)) %>%
  group_by(Pseudotime_bin) %>%
  mutate(Rank_bin = rank(Score)) %>%
  pivot_wider(id_cols = c("Gene"), names_from = Pseudotime_bin, values_from = c(Score, Rank_bin)) %>%
  rowwise() %>% # take mean across row of ranks
  mutate(Rank_mean = mean(c(Rank_bin_5, Rank_bin_6, Rank_bin_7), na.rm = TRUE)) %>%
  rowwise() %>%
  mutate(Score_mean = mean(c(Score_5, Score_6, Score_7), na.rm = TRUE)) %>%
  arrange(Rank_mean, Score_mean) 

TF_CD8_table
```

## Combine results

```{r CD4 megatable}
CD4_megatable <- TF_CD4_table %>%
  select(Gene, contains("Rank_bin"), contains("Score_"), Score_mean, Rank_mean) %>% 
  mutate(TCR_pathway = Gene %in% TCR_set) %>%
  mutate(Targets_masterreg = Gene %in% Targets_masterreg) %>%
  left_join(DE_CD4_table, by = c("Gene")) %>%
  pivot_wider(values_from = LFC, names_from = clusters, names_prefix = "LFC_bin_")
```

```{r CD8 megatable}
CD8_megatable <- TF_CD8_table %>%
  select(Gene, contains("Rank_bin"), contains("Score_"), Score_mean, Rank_mean) %>%
  mutate(TCR_pathway = Gene %in% TCR_set) %>%
  mutate(Targets_masterreg = Gene %in% Targets_masterreg) %>%
  left_join(DE_CD8_table, by = c("Gene")) %>%
  pivot_wider(values_from = LFC, names_from = clusters, names_prefix = "LFC_bin_")
```

## Plot

```{r Plot CD4 megatable}
TF_number = 50

# Use TCR and is_DE as annotation
CD4_megatable <- CD4_megatable %>%
  arrange(Rank_mean, Score_mean)
CD4_mat <- as.matrix(CD4_megatable[1:TF_number, c("Rank_bin_4", "Rank_bin_5", "Rank_bin_6", "Rank_bin_7")])
row.names(CD4_mat) <- CD4_megatable[1:TF_number, "Gene"] %>% pull()

# Annotation
CD4_annotation <- (as.data.frame(CD4_megatable[1:TF_number, c("TCR_pathway", "Targets_masterreg")]))
colnames(CD4_annotation) <- c("TCR pathway", "Targets master regulator")
CD4_annotation$`TCR pathway` <- factor(CD4_annotation$`TCR pathway`, labels = c("No", "Yes"))
CD4_annotation$`Targets master regulator` <- factor(CD4_annotation$`Targets master regulator`, labels = c("No", "Yes"))
row.names(CD4_annotation) <- CD4_megatable[1:TF_number, "Gene"] %>% pull()

# Add DE results to annotations (is it DE in any of the relevant time bins?)
CD4_megatable <- CD4_megatable %>%
  rowwise() %>%
  mutate("Differentially expressed" = !(all(is.na((c(LFC_bin_4, LFC_bin_5, LFC_bin_6, LFC_bin_7))))))
CD4_megatable$`Differentially expressed` <- factor(CD4_megatable$`Differentially expressed`, labels = c("No", "Yes"))
CD4_annotation$`Differentially expressed` <- CD4_megatable[1:TF_number, "Differentially expressed"] %>% pull()

annotation_colors <-  list(
  `TCR pathway` = c(No = "white", Yes = "darkgreen"),
  `Targets master regulator` = c(No = "white", Yes = "darkorchid4"),
  `Differentially expressed` = c(No = "white", Yes = "darkred")
  )

pheatmap((CD4_mat),
         cluster_cols = FALSE,
         cluster_rows = FALSE,
         show_colnames= TRUE,
         show_rownames = TRUE,
         cellwidth = 10,
         cellheight = 10,
         angle_col = 90,
         annotation_row = (CD4_annotation),
         labels_col = c("Pseudotime 4-5", "Pseudotime 5-6", "Pseudotime 6-7", "Pseudotime 7-8"),
         color = colorRampPalette(brewer_pal(palette = "Blues", direction = -1)(9), bias = 2)(100), # gray na_col for NA values (< 5% cells express)
         annotation_colors = annotation_colors,
         #filename = str_c("figures/", "CD4_megatable_50gene", ".pdf"),
         height = 10,
         width = 5
         )
```

```{r CD8 plot megatable}
TF_number = 50

# Use TCR and is_DE as annotation
CD8_megatable <- CD8_megatable %>%
  arrange(Rank_mean, Score_mean)
CD8_mat <- as.matrix(CD8_megatable[1:TF_number, c("Rank_bin_5", "Rank_bin_6", "Rank_bin_7", "Rank_bin_8")])
row.names(CD8_mat) <- CD8_megatable[1:TF_number, "Gene"] %>% pull()

# Annotation
CD8_annotation <- (as.data.frame(CD8_megatable[1:TF_number, c("TCR_pathway", "Targets_masterreg")]))
colnames(CD8_annotation) <- c("TCR pathway", "Targets master regulator")
CD8_annotation$`TCR pathway` <- factor(CD8_annotation$`TCR pathway`, labels = c("No", "Yes"))
CD8_annotation$`Targets master regulator` <- factor(CD8_annotation$`Targets master regulator`, labels = c("No", "Yes"))
row.names(CD8_annotation) <- CD8_megatable[1:TF_number, "Gene"] %>% pull()

# Add DE results to annotations (is it DE in any of the relevant time bins?)
CD8_megatable <- CD8_megatable %>%
  rowwise() %>%
  mutate("Differentially expressed" = !(all(is.na((c(LFC_bin_5, LFC_bin_6, LFC_bin_7, LFC_bin_8))))))
CD8_megatable$`Differentially expressed` <- factor(CD8_megatable$`Differentially expressed`, labels = c("No", "Yes"))
CD8_annotation$`Differentially expressed` <- CD8_megatable[1:TF_number, "Differentially expressed"] %>% pull()

annotation_colors <-  list(
  `TCR pathway` = c(No = "white", Yes = "darkgreen"),
  `Targets master regulator` = c(No = "white", Yes = "darkorchid4"),
  `Differentially expressed` = c(No = "white", Yes = "darkred")
  )

pheatmap((CD8_mat),
         cluster_cols = FALSE,
         cluster_rows = FALSE,
         show_colnames= TRUE,
         show_rownames = TRUE,
         cellwidth = 10,
         cellheight = 10,
         angle_col = 90,
         annotation_row = (CD8_annotation),
         labels_col = c("Pseudotime 5-6", "Pseudotime 6-7", "Pseudotime 7-8", "Pseudotime 8-9"),
         color = colorRampPalette(brewer_pal(palette = "Blues", direction = -1)(9), bias = 2)(100),
         annotation_colors = annotation_colors,
         #filename = str_c("figures/", "CD8_megatable_50gene", ".pdf"),
         height = 10,
         width = 5 
         )
```

# Plot DE results as heatmap

```{r Find cluster order}
# CD4 order: Find average peak fold change  of clusters
cd4_DE_heat <- DE_smallbins_lin %>%
  arrange(clusters, desc(lfc_median)) %>%
  inner_join(cd4_gene_clusters %>% select(Gene_cluster = Cluster, everything()), by = "Gene") %>%
  select(Gene, lfc_median, clusters, Gene_cluster) %>%
  pivot_wider(values_from = lfc_median, names_from = clusters, names_prefix = "Pseudotime_") %>%
  arrange(Gene_cluster) %>%
  
  # Find peak and arrange
  rowwise() %>%
  mutate(peak_lfc = which.max(c_across(Pseudotime_0:Pseudotime_14))) %>% # find max per row
  group_by(Gene_cluster) %>% # group by gene cluster
  summarize(cluster_peak = mean(peak_lfc)) %>% # get mean peak per cluster
  arrange(cluster_peak)

# Get cluster order:
cd4_cluster_order <- factor(cd4_DE_heat %>% select(Gene_cluster) %>% pull(), levels = cd4_DE_heat %>% select(Gene_cluster) %>% pull())

# CD8 order: find average peak fold change  of clusters
cd8_DE_heat <- DE_smallbins_lin %>%
  arrange(clusters, lfc_median) %>%
  inner_join(cd8_gene_clusters %>% select(Gene_cluster = Cluster, everything()), by = "Gene") %>%
  select(Gene, lfc_median, clusters, Gene_cluster) %>%
  pivot_wider(values_from = lfc_median, names_from = clusters, names_prefix = "Pseudotime_") %>%
  arrange(Gene_cluster) %>%
  # find peak and arrange
  rowwise() %>%
  mutate(peak_lfc = which.min(c_across(Pseudotime_0:Pseudotime_14))) %>% # find min per row
  group_by(Gene_cluster) %>% # group by gene cluster
  summarize(cluster_peak = mean(peak_lfc)) %>% # get mean peak per cluster
  arrange((cluster_peak))

# Get cluster order:
cd8_cluster_order <- factor(cd8_DE_heat %>% select(Gene_cluster) %>% pull(), levels = cd8_DE_heat %>% select(Gene_cluster) %>% pull())
```

```{r Select DE rows to annotate}
# For annotation
cd4_row_labels <- c("Cd4", "Id2", "Cd5", "Gata3", "Zbtb7b", "Egr2", "Cd69") # "Egr1", "Cd40lg", 
cd8_row_labels <- c("Cd8a", "Themis", "Runx3", "Nkg7", "Egr2", "Cd69", "Kcna2", "Tmie") # "Egr1", 

cd4_labels_de <- DE_smallbins_lin %>%
  arrange(clusters, (lfc_median)) %>%
  inner_join(cd4_gene_clusters %>% select(Gene_cluster = Cluster, everything()), by = "Gene") %>%
  select(Gene, lfc_median, clusters, Gene_cluster) %>%
  pivot_wider(values_from = lfc_median, names_from = clusters, names_prefix = "Pseudotime_") %>%
  mutate(Gene_cluster = factor(Gene_cluster, levels = cd4_cluster_order)) %>% # make gene cluster a factor that is ordered by the above levels
  arrange(Gene_cluster) %>%
  select(Gene) %>%
  mutate(row_labels = case_when(Gene %in% cd4_row_labels ~ Gene,
                                TRUE ~ "")) %>%
  select(row_labels) %>%
  pull()

cd8_labels_de <- DE_smallbins_lin %>%
  arrange(clusters, desc(lfc_median)) %>%
  inner_join(cd8_gene_clusters %>% select(Gene_cluster = Cluster, everything()), by = "Gene") %>%
  select(Gene, lfc_median, clusters, Gene_cluster) %>%
  pivot_wider(values_from = lfc_median, names_from = clusters, names_prefix = "Pseudotime_") %>%
  mutate(Gene_cluster = factor(Gene_cluster, levels = cd8_cluster_order)) %>% # make gene cluster a factor that is ordered by the above levels
  arrange(Gene_cluster) %>%
  select(Gene) %>%
  mutate(row_labels = case_when(Gene %in% cd8_row_labels ~ Gene,
                                TRUE ~ "")) %>%
  select(row_labels) %>%
  pull()
```

```{r CD4 DE heatmap}
bin_ranges_labels <- c(`0` = "0-1",
                             `1` = "1-2",
                             `2` = "2-3",
                             `3` = "3-4",
                             `4` = "4-5",
                             `5` = "5-6",
                             `6` = "6-7",
                             `7` = "7-8",
                             `8` = "8-9",
                             `9` = "9-10",
                             `10` = "10-11",
                             `11` = "11-12",
                             `12` = "12-13",
                             `13` = "13-14",
                             `14` = "14-16")

# Collect DE data
cd4_DE_heat <- DE_smallbins_lin %>%
  arrange(clusters, (lfc_median)) %>%
  inner_join(cd4_gene_clusters %>% select(Gene_cluster = Cluster, everything()), by = "Gene") %>%
  select(Gene, lfc_median, clusters, Gene_cluster) %>%
  pivot_wider(values_from = lfc_median, names_from = clusters, names_prefix = "Pseudotime_") %>%
  mutate(Gene_cluster = factor(Gene_cluster, levels = cd4_cluster_order)) %>% # make gene cluster a factor that is ordered by the above levels
  arrange(Gene_cluster)

cd4_DE_heatmat <- as.matrix(cd4_DE_heat %>% select(contains("Pseudotime")))
row.names(cd4_DE_heatmat) <- cd4_DE_heat$Gene

cd4_heatmat_labels <- as.data.frame(cd4_DE_heat$Gene_cluster)
row.names(cd4_heatmat_labels) <- cd4_DE_heat$Gene
colnames(cd4_heatmat_labels) <- "Gene cluster"

annotation_colors_cd4de <- list("Gene cluster" = sample(stepped2(20)[10:18]))
names(annotation_colors_cd4de[[1]]) <- as.character(seq(0,8))

# PLot LFCs as a heatmap
pheatmap(cd4_DE_heatmat,
         cluster_cols = FALSE,
         cluster_rows = FALSE,
         show_colnames= TRUE,
         show_rownames = TRUE,
         cellwidth = 10,
         angle_col = 90,
         labels_col = bin_ranges_labels,
         annotation_row = cd4_heatmat_labels,
         labels_row = cd4_labels_de,
         color = colorRampPalette(brewer_pal(palette = "RdBu", direction = -1)(9))(100),
         annotation_colors = annotation_colors_cd4de,
         main = "MHCII-specific DE genes",
         annotation_names_row = FALSE,
         breaks = c(seq(.9*min(cd4_DE_heatmat), 0, length.out=51), 
              seq(max(cd4_DE_heatmat)/50, .9*max(cd4_DE_heatmat), length.out=50)), # set the 0 point for color, see https://stackoverflow.com/questions/31677923/set-0-point-for-pheatmap-in-r
         #filename = str_c("figures/", "cd4_de_heatmap", ".pdf"),
         height = 6,
         width = 4.5,
         fontsize_row = 10
         )
```

```{r CD8 DE heatmap}
# Collect DE data
cd8_DE_heat <- DE_smallbins_lin %>%
  arrange(clusters, desc(lfc_median)) %>%
  inner_join(cd8_gene_clusters %>% select(Gene_cluster = Cluster, everything()), by = "Gene") %>%
  select(Gene, lfc_median, clusters, Gene_cluster) %>%
  pivot_wider(values_from = lfc_median, names_from = clusters, names_prefix = "Pseudotime_") %>%
  mutate(Gene_cluster = factor(Gene_cluster, levels = cd8_cluster_order)) %>% # make gene cluster a factor that is ordered by the above levels
  arrange(Gene_cluster)

cd8_DE_heatmat <- as.matrix(cd8_DE_heat %>% select(contains("Pseudotime")))
row.names(cd8_DE_heatmat) <- cd8_DE_heat$Gene

cd8_heatmat_labels <- as.data.frame(cd8_DE_heat$Gene_cluster)
row.names(cd8_heatmat_labels) <- cd8_DE_heat$Gene
colnames(cd8_heatmat_labels) <- "Gene cluster"

annotation_colors_cd8de <- list("Gene cluster" =  sample(stepped2(20)[1:8]))
names(annotation_colors_cd8de[[1]]) <- as.character(seq(0,7))

# PLot LFCs as a heatmap
pheatmap(cd8_DE_heatmat,
         cluster_cols = FALSE,
         cluster_rows = FALSE,
         show_colnames= TRUE,
         show_rownames = TRUE,
         cellwidth = 10,
         angle_col = 90,
         labels_col = bin_ranges_labels, 
         annotation_row = cd8_heatmat_labels,
         labels_row = cd8_labels_de, 
         color = colorRampPalette(brewer_pal(palette = "RdBu", direction = -1)(9))(100), 
         annotation_colors = annotation_colors_cd8de,
         main = "MHCI-specific DE genes",
         annotation_names_row = FALSE,
         breaks = c(seq(.9*min(cd8_DE_heatmat), 0, length.out=51), 
              seq(max(cd8_DE_heatmat)/50, .9*max(cd8_DE_heatmat), length.out=50)), 
         #filename = str_c("figures/", "cd8_de_heatmap", ".pdf"),
         height = 6,
         width = 4.5,
         fontsize_row = 10
         )
```
# Enrichment of TCR signaling in gene clusters

```{r Hypergeometric test for CD4 clusters}
cd4_gene_clusters$Cluster <- as.factor(cd4_gene_clusters$Cluster)

p_values <- c()
for (cluster in levels(cd4_gene_clusters$Cluster)) {
  DE_geneset <- cd4_gene_clusters %>%
    filter(Cluster == cluster) %>%
    select(Gene) %>%
    pull()
  overlap <- length(intersect(DE_geneset, TCR_sig))
  hyper_out <- phyper(q = overlap - 1, m = length(TCR_sig), n = length(totalVI_genes) - length(TCR_sig), k = length(DE_geneset),
       lower.tail = FALSE, log.p = FALSE)
  p_values <- c(p_values, hyper_out)
  print(str_c("Cluster: ", cluster, ", p-value: ", hyper_out))
}

p_values_adjusted <- p.adjust(p_values, method = "BH")
names(p_values_adjusted) <- levels(cd4_gene_clusters$Cluster)
p_values_adjusted
```

```{r Hypergeometric test for CD8 clusters}
cd8_gene_clusters$Cluster <- as.factor(cd8_gene_clusters$Cluster)

p_values <- c()
for (cluster in levels(cd8_gene_clusters$Cluster)) {
  DE_geneset <- cd8_gene_clusters %>%
    filter(Cluster == cluster) %>%
    select(Gene) %>%
    pull()
  overlap <- length(intersect(DE_geneset, TCR_sig))
  hyper_out <- phyper(q = overlap - 1, m = length(TCR_sig), n = length(totalVI_genes) - length(TCR_sig), k = length(DE_geneset),
       lower.tail = FALSE, log.p = FALSE)
  p_values <- c(p_values, hyper_out)
  print(str_c("Cluster: ", cluster, ", p-value: ", hyper_out))
}

p_values_adjusted <- p.adjust(p_values, method = "BH")
names(p_values_adjusted) <- levels(cd8_gene_clusters$Cluster)
p_values_adjusted
```

# Stat5 signature

```{r Load Stat5 signature}
hallmark_sig_file <- "/data/yosef/users/zsteier/references/Signatures/h.all.v7.0.symbols.gmt"
# https://www.gsea-msigdb.org/gsea/msigdb/mouse/geneset/HALLMARK_IL2_STAT5_SIGNALING.html

sig_string <- "Cd86,Traf1,Pou2f1,Ecm1,Serpinb6a,Cish,Fgl2,Tiam1,Tnfrsf4,Cdkn1c,Penk,Rora,Gpx4,Ltb,Ahr,Slc1a5,Il4ra,Ptch1,Adam19,Il18r1,Nrp1,Myo1c,Myo1e,Tnfsf10,Capn3,Scn9a,Ifngr1,Casp3,Gadd45b,Ccr4,Nop2,Lrig1,Emp1,Rhob,Gpr65,Rgs16,Plpp1,Ncs1,Mapkapk2,Nfil3,Bmpr2,Cd81,Irf4,Phlda1,Bhlhe40,Capg,Amacr,Rragd,Hycc2,She,Tnfsf11,Plagl1,Tnfrsf9,Rnh1,Dennd5a,Eomes,Map6,Socs2,Ncoa3,Plec,Coch,F2rl2,Cst7,Itgae,Umps,Swap70,Alcam,Hipk2,Tnfrsf1b,Hk2,Dhrs3,Ahnak,St3gal4,Cd83,Syngr2,Cyfip1,P2rx4,S100a1,Csf2,Csf1,Ndrg1,Gsto1,Ikzf2,Ikzf4,Spry4,Cdc6,Slc29a2,Klf6,Map3k8,Cxcl10,Rabgap1l,Socs1,Icos,Batf,Irf6,Syt11,Praf2,Gucy1b1,Ctsz,Ifitm3,Snx9,Slc39a8,Gabarapl1,Eef1akmt1,Pdcd2l,Sh3bgrl2,Wls,Phtf2,Dcps,Hopx,Ttc39b,Glipr2,Cdc42se2,Rhoh,Batf3,Itih5,Gbp3,Huwe1,Pus1,Smpdl3a,Nfkbiz,Uck2,Twsg1,Lrrc8c,Spred2,Tnfrsf21,Snx14,Tlr7,Cdcp1,Galm,Etfbkmt,Ptrh2,Ckap4,Lclat1,Drc1,Ahcyl,Plin2,Anxa4,Aplp1,Serpinc1,Bcl2,Bcl2l1,Bmp2,Car2,Ccnd2,Ccnd3,Ccne1,Cd44,Cd48,Col6a1,Ctla4,Plscr1,Ager,Tnfrsf18,Eno3,Fah,Flt3l,Gata1,Gpr83,Slc2a3,Irf8,Cd79b,Igf1r,Igf2r,Il10,Il10ra,Il13,Il1r2,Il2ra,Il2rb,Il3ra,Itga6,Itgav,Lif,Mxd1,Maff,Muc1,Myc,Pnp,Enpp1,Odc1,P4ha1,Furin,Abcb1a,Pim1,Prkch,Prnp,Ptger2,Pth1r,Sell,Selp,Spp1,Il1rl1,Tgm2,Xbp1,Etv4,Arl4a,Nt5e,Tnfrsf8"
sig_list <- str_split(sig_string, ",")
```

```{r Plot average expression of signature}
Stat5_genes_annotated <- denoised_genes_annotated %>%
  filter(Gene %in% sig_list[[1]]) %>%
  group_by(mean_pseudotime, Lineage_by_genotypeSlingshot) %>% ## group by Cluster, time, lineage and take mean
  summarize(sig_mean = mean(value))

# Plot by loess, facet by cluster
ggplot(Stat5_genes_annotated, aes(x = mean_pseudotime, y = sig_mean, color = Lineage_by_genotypeSlingshot)) +
  geom_smooth(method = "loess", span = .5, se = FALSE) +
  scale_color_manual(values = c(redcol, bluecol)) +
  labs(x = "Pseudotime", y = "Mean scaled\n denoised expression", color = "Lineage", title = "Hallmark Il2-Stat5 Signaling") +
  theme_classic()
ggsave("figures/Stat5_signature_pseudotime.pdf", bg = "transparent", height = 2, width = 4)
```